---
title: "Annotating footprints using motif cluster"
author: "XSun"
date: "2026-03-12"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r message=FALSE, warning=FALSE}

library(ggplot2)
library(data.table)
library(patchwork)

#fp_lengths <- c("10-30bp", "30-60bp", "60-80bp")
fp_lengths <- c("60-80bp")
fp_length <- "60-80bp"

base_out_dir     <- "/project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation"
```


# Introduction

This analysis aimed to annotate Fiber-seq footprints with known motif clusters from the Vierstra motif clustering resource (`/project/spott/kevinluo/Fiber_seq/data/motifs/Vierstra_motif_clustering/hg38.archetype_motifs.v1.0.bed.gz`, collected by Kevin).  

The workflow consisted of two major stages:

1. **Genome-wide overlap of footprint intervals with motif matches** using `bedtools`.

  `/project/spott/xsun/fiberseq/2.annotating_footprints/13.motif_cluster_by_chr.sbatch`
  `/project/spott/xsun/fiberseq/2.annotating_footprints/14.fp_overlap_motifcluster.sbatch`
  `/project/spott/xsun/fiberseq/2.annotating_footprints/15.run_fp_motif_overlap_by_chr.sbatch`
  `/project/spott/xsun/fiberseq/2.annotating_footprints/16.submit_fp_motif_overlap_jobs.sh`


2. **Downstream summarization and visualization in R**, including motif assignment per footprint and descriptive statistics across footprint size bins.

The footprint size bins analyzed were:

- `10-30bp`
- `30-60bp`
- `60-80bp`

To keep only reasonably motif matches, overlaps were filtered by requiring at least 5 bp overlap:

# Results

The processed overlaps are storaged here: `/project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/`

Taking the footprints in `60-80bp` as an example

- `annotate_fp_motif.filtered_overlap.RDS` & `annotate_fp_motif.filtered_overlap.tsv.gz`:  full filtered overlap table after applying overlap_bp >= min_overlap (5bp) 

```{bash }

zcat /project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/60-80bp/annotate_fp_motif.filtered_overlap.tsv.gz | head

```

- `annotate_fp_motif.top1.RDS`  &  `annotate_fp_motif.top1.tsv.gz`    : keep only the best motif per footprint (based on `match_score`).      

```{bash }

zcat /project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/60-80bp/annotate_fp_motif.top1.tsv.gz | head

```

- `annotate_fp_motif.top5.RDS`  &  `annotate_fp_motif.top5.tsv.gz`    : keep top 5 motifs per footprint (based on `match_score`).         

```{bash }

zcat /project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/60-80bp/annotate_fp_motif.top5.tsv.gz | head

```

## How many footprints can be annotated?

A footprint is considered annotated if it has at least one motif passing the overlap filter.

```{r fig.height=4, fig.width=12}

p <- list()
for (fp_length in fp_lengths){
  
  annotation_summary <- readRDS(paste0("/project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/", fp_length, "/annotation_summary.RDS"))
  
  df_plot <- data.table(
    category = c("Annotated footprints", "Unannotated footprints"),
    count = c(annotation_summary$annotated_footprints, annotation_summary$unannotated_footprints)
  )
  df_plot[, percent := count / sum(count) * 100]
  
  p[[length(p)+1]] <- ggplot(df_plot, aes(x = category, y = count, fill = category)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", percent)),
            vjust = -0.4, size = 5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_classic(base_size = 14) +
  labs(
    x = "",
    y = "Number of footprints",
    title = paste0("Footprint annotation: ", fp_length)
  ) +
  theme(legend.position = "none")
  
}

wrap_plots(p)

```

## Do footprints usually have one motif or multiple motifs?

This is summarized by counting how many motif hits are associated with each footprint. 

```{r fig.height=4, fig.width=12}


p <- list()
for (fp_length in fp_lengths){
  
  fp_dist_summary <- readRDS(paste0("/project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/", fp_length, "/motifs_per_footprint_summary.RDS"))
  
  plot_dt <- data.table(
    category = c("1 motif", "Multiple motifs"),
    count = c(
      fp_dist_summary$footprints_with_1_motif,
      fp_dist_summary$footprints_with_multiple_motifs
    ),
    proportion = c(
      fp_dist_summary$prop_with_1_motif,
      fp_dist_summary$prop_with_multiple_motifs
    )
  )
  
  
  p[[length(p)+1]] <- ggplot(plot_dt, aes(x = category, y = count, fill = category)) +
    geom_col(width = 0.6) +
    geom_text(
      aes(label = sprintf("%d\n(%.1f%%)", count, proportion * 100)),
      vjust = -0.3,
      size = 5
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_classic(base_size = 14) +
    labs(
      x = "",
      y = "Number of annotated footprints",
      title = paste0("Motif multiplicity per footprint: ", fp_dist_summary$fp_length)
    ) +
    theme(legend.position = "none")
  
  
}

wrap_plots(p)

p <- list()
for (fp_length in fp_lengths){
  
  distribution <- readRDS(paste0("/project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/", fp_length, "/motifs_per_footprint_distribution.RDS"))

  p[[length(p)+1]] <- ggplot(distribution, aes(x = n_motifs, y = N)) +
    geom_col(width = 1, fill = "#4C72B0") +
    scale_y_log10() +
    theme_classic(base_size = 14) +
    labs(
      x = "Number of motifs per footprint",
      y = "Number of footprints (log scale)",
      title = paste0("Distribution of motifs per footprint: ", fp_dist_summary$fp_length)
    )
  
  
}

wrap_plots(p)

```



## What motifs have the most footprints?

For each motif cluster, the number of unique footprints containing that motif is counted.

```{r fig.height=12, fig.width=12}

top_n <- 20

p <- list()
for (fp_length in fp_lengths){
  
  motif_fp_counts <- readRDS(paste0("/project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/", fp_length, "/motif_footprint_counts.RDS"))
  
  df_plot <- motif_fp_counts[order(-n_footprints)][1:top_n]
  
  p[[length(p)+1]] <- ggplot(df_plot,
                             aes(x = reorder(motif_cluster, n_footprints), y = n_footprints)) +
    geom_col(width = 0.8) +
    coord_flip() +
    theme_classic(base_size = 14) +
    labs(
      x = "Motif cluster",
      y = "Number of footprints",
      title = paste0("Top ", top_n, " motif clusters by annotated footprints: ", fp_length)
    )
  
  
  
  
}

wrap_plots(p, ncol = 1)

```



## For each motif, how many footprints are found?

For each motif cluster, the number of unique footprints containing that motif is counted:


<!-- ```{r fig.height=4, fig.width=12} -->

<!-- motif_fp_counts <- readRDS(paste0("/project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/10-30bp/motif_footprint_counts.RDS")) -->

<!-- DT::datatable(motif_fp_counts,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','10-30bp'),options = list(pageLength = 10) ) -->
<!-- ``` -->

<!-- ```{r fig.height=4, fig.width=12} -->

<!-- motif_fp_counts <- readRDS(paste0("/project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/30-60bp/motif_footprint_counts.RDS")) -->

<!-- DT::datatable(motif_fp_counts,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','30-60bp'),options = list(pageLength = 10) ) -->
<!-- ``` -->


```{r fig.height=4, fig.width=12}

motif_fp_counts <- readRDS(paste0("/project/spott/xsun/fiberseq/2.annotating_footprints/results/annotate_fp_motif/motif_annotation/60-80bp/motif_footprint_counts.RDS"))

DT::datatable(motif_fp_counts,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','60-80bp'),options = list(pageLength = 10) )
```
