---
title: "Checking the cCRE peak coverage for 15 states from ChromHMM, GM12878"
author: "XSun"
date: "2026-02-25"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r}

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
})

```


# Introduction

We checked cCREs peak from GM12878 overlap with 15 states from ChromHMM, to calibrate the FIRE calling

# Data overview

paper: https://www.nature.com/articles/s41586-025-09909-9

ENCODE cCREs from SCREEN https://screen.wenglab.org/downloads

`/project/spott/kevinluo/Fiber_seq/data/ENCODE_CREs/cCREs_v4`

`GRCh38-cCREs.bed` : includes everything and has a column with the type of cCREs, it is a union set of 1,888 cell types

A subdirectory named `LCLs/` contains cCRE annotations specific to several lymphoblastoid cell lines (LCLs):  Some cell lines (e.g., GM12878) integrate multiple data types (DNase, ATAC, histone marks, CTCF), whereas others (e.g., GM19238) are primarily supported by DNase data.

- GM12865
- GM12878
- GM12891
- GM19238

We use the data from GM12878 here.


```{r}

peak_file <- "/project/spott/kevinluo/Fiber_seq/data/ENCODE_CREs/cCREs_v4/LCLs/GM12878_cCREs_ENCFF428XFI_ENCFF280PUF_ENCFF469WVA_ENCFF644EEX.bed"
peak_df <- fread(peak_file, showProgress = FALSE)

DT::datatable(head(peak_df),caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','cCRE data structure, GM12878'),options = list(pageLength = 10) )

peak_df[, width := as.numeric(V3 - V2)]

# -----------------------------
# Detailed category + description
# -----------------------------
peak_df[, `:=`(
  cCRE_group = fcase(
    V10 == "PLS",         "PLS",
    V10 == "pELS",        "pELS",
    V10 == "dELS",        "dELS",
    V10 == "CA-only",     "CA-only",
    V10 == "TF",          "TF",
    V10 == "CA-TF",       "CA-TF",
    V10 == "CA-CTCF",     "CA-CTCF",
    V10 == "CA-H3K4me3",  "CA-H3K4me3",
    V10 == "Low-DNase", "Low-DNase", 
    default = "Other"
  ),
  cCRE_desc = fcase(
    V10 == "PLS",         "Promoter-like signature",
    V10 == "pELS",        "Proximal enhancer-like signature",
    V10 == "dELS",        "Distal enhancer-like signature",
    V10 == "CA-only",     "Chromatin accessible only",
    V10 == "TF",          "TF only",
    V10 == "CA-TF",       "Chromatin accessible with TF",
    V10 == "CA-CTCF",     "Chromatin accessible with CTCF",
    V10 == "CA-H3K4me3",  "Chromatin accessible with H3K4me3",
    V10 == "Low-DNase", "Low-DNase", 
    default = "Other / unknown"
  )
)]

# -----------------------------
# Summary table (by detailed category)
# -----------------------------
summary_table <- peak_df[, .(
  n_peaks        = .N,
  mean_width     = round(mean(width), 1),
  median_width   = round(median(width), 1),
  total_bp       = sum(width)
), by = .(cCRE_group, cCRE_desc)][order(-n_peaks)]



DT::datatable(summary_table,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','cCRE overview, GM12878 '),options = list(pageLength = 10) )

```




# Overlap with ChromHMM 


```{r fig.height=6, fig.width=7}

res_allcat <- readRDS("/project/spott/xsun/fiberseq/2.annotating_footprints/results/cCRE_ChromHMM_GM12878/cCRE_all_and_byCategory_vs_ChromHMM.rds")

for (label in unique(res_allcat$peak_label)) {
  
  res <- as.data.table(copy(res_allcat[res_allcat$peak_label == label]))
  
  # Keep/standardize columns
  res <- res[, .(state_id, description, hex,
                 frac_state_covered_by_peak, frac_of_peak_bp_in_state)]
  
  # Convert to percent
  res[, `%state_covered_by_peak` := frac_state_covered_by_peak * 100]
  res[, `%peak_in_state` := frac_of_peak_bp_in_state * 100]
  
  # Ensure ordering by state_id
  setorder(res, state_id)
  
  # -------------------------
  # Bar plot (two metrics stacked)
  # -------------------------
  plot_long <- melt(
    res[, .(state_id, description, hex, `%state_covered_by_peak`, `%peak_in_state`)],
    id.vars = c("state_id", "description", "hex"),
    measure.vars = c("%state_covered_by_peak", "%peak_in_state"),
    variable.name = "metric",
    value.name = "value"
  )
  
  plot_long[, metric := fifelse(
    metric == "%state_covered_by_peak",
    "% of state covered by peak",
    "% of peak in state"
  )]
  
  p_bar <- ggplot(plot_long, aes(x = state_id, y = value, fill = hex)) +
    geom_col() +
    facet_wrap(~ metric, ncol = 1, scales = "free_y") +
    scale_fill_identity() +
    scale_x_continuous(
      breaks = res$state_id,
      labels = res$description
    ) +
    labs(
      x = "ChromHMM state",
      y = "Percent (%)",
      title = paste0(label, ": cCRE & ChromHMM overlap")
    ) +
    theme_classic(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold"),
      legend.position = "none"
    )
  
  # -------------------------
  # Pie chart (%peak in state)
  # -------------------------
  pie_df <- copy(res)
  
  # If no overlap at all, `%peak_in_state` can be NA -> set to 0 to avoid errors
  pie_df[is.na(`%peak_in_state`), `%peak_in_state` := 0]
  
  pie_df[, state_label := as.character(state_id)]
  pie_df[, legend_label := paste0(state_id, ": ", description)]
  pie_df[, legend_label := factor(legend_label, levels = legend_label)]
  
  p_pie <- ggplot(pie_df,
                  aes(x = "", y = `%peak_in_state`, fill = legend_label)) +
    geom_col(width = 1, color = "white") +
    coord_polar(theta = "y") +
    scale_fill_manual(values = setNames(pie_df$hex, pie_df$legend_label)) +
    theme_void(base_size = 14) +
    labs(
      title = paste0(label, ": % of peak in each ChromHMM state"),
      fill = "ChromHMM States"
    ) +
    geom_text(
      aes(label = state_label),
      position = position_stack(vjust = 0.5),
      size = 4,
      color = "black"
    )

  print(p_bar)
  print(p_pie)
}



```
