---
title: "Checking the DNase peak coverage for 15 states from ChromHMM"
author: "XSun"
date: "2026-02-24"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

# Introduction

We checked DNase peak overlap with 15 states from ChromHMM, to calibrate the FIRE calling

The results for FIRE peak is here: https://xsun1229.github.io/fiberseq/fire_coverage_ChromHMM.html

Processed DNase-seq peak files (BED format, `.bed.gz`) were downloaded from the ENCODE Consortium by Kevin and stored locally at: `/project/spott/kevinluo/Fiber_seq/data/DNaseseq/`

Cell lines included: currently include four LCLs:

- GM12865
- GM12878
- GM12891
- GM19238

GM19238 is in our Fiber-seq sample list


**We sum the total number of base pairs of DNase peaks inside each of the 15 ChromHMM states , and then divided by the total number of base pairs for each states.**


```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(IRanges)
  library(data.table)
  library(ggplot2)
})
```


# Results


```{r fig.height=6, fig.width=7}

cellines <- c("GM19238","GM12865", "GM12878", "GM12891")

folder_results <- "/project/spott/xsun/fiberseq/2.annotating_footprints/results/DNase/"

for (celline in cellines){
  
  res <- readRDS(paste0(folder_results,"any_overlap_with_ChromHMM_", celline, ".RDS"))
  
  res <- res[,-c("total_bp_state","peak_overlap_bp")]
  res <- res[,c("state_id","description","total_state_Mb","peak_overlap_Mb","frac_state_covered_by_peak","frac_of_peak_bp_in_state","hex")]
  res$frac_state_covered_by_peak <- res$frac_state_covered_by_peak*100
  res$frac_of_peak_bp_in_state <- res$frac_of_peak_bp_in_state *100
  colnames(res)[5:6] <- c("%state_covered_by_peak","%peak_in_state")
  
  
  plot_df <- copy(res)
  
  plot_long <- melt(
    plot_df,
    id.vars = c("state_id", "description", "hex"),
    measure.vars = c("%state_covered_by_peak", "%peak_in_state"),
    variable.name = "metric",
    value.name = "value"
  )
  
  # Simple labels
  plot_long[, metric := fifelse(
    metric == "%state_covered_by_peak",
    "% of state covered by peak",
    "% of peak in state"
  )]
  
  plot <- ggplot(plot_long, aes(x = state_id, y = value, fill = hex)) +
    geom_col() +
    facet_wrap(~ metric, ncol = 1, scales = "free_y") +
    scale_fill_identity() +
    scale_x_continuous(
      breaks = plot_df$state_id,
      labels = plot_df$description
    ) +
    labs(
      x = "ChromHMM state",
      y = "Percent (%)",
      title = paste0(celline,": DNase peak & ChromHMM Overlap")
    ) +
    theme_classic(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold"),
      legend.position = "none"
    )
  
  print(plot)
  
  
  pie_df <- copy(res)
  
  # Ensure ordering by state_id
  setorder(pie_df, state_id)
  
  # Create numeric labels (1–15)
  pie_df[, state_label := as.character(state_id)]
  
  # Create legend labels: "1: Active TSS"
  pie_df[, legend_label := paste0(state_id, ": ", description)]
  pie_df$legend_label <- factor(pie_df$legend_label)
  
  pie <- ggplot(pie_df,
         aes(x = "", 
             y = `%peak_in_state`,
             fill = legend_label)) +
    geom_col(width = 1, color = "white") +
    coord_polar(theta = "y") +
    scale_fill_manual(values = setNames(pie_df$hex,
                                        pie_df$legend_label)) +
    theme_void(base_size = 14) +
    labs(
      title = paste0(celline,": % of peak in each ChromHMM state"),
      fill = "ChromHMM States"
    ) +
    geom_text(
      aes(label = state_label),
      position = position_stack(vjust = 0.5),
      size = 4,
      color = "black"
    )
  
  print(pie)
  
}


```






