---
title: "Overlap of DNase peaks with cCREs peaks"
author: "XSun"
date: "2026-03-10"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---


```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(GenomicRanges)
  library(IRanges)
  library(ggplot2)
  library(patchwork)
})

keep_chr <- paste0("chr", 1:22)


#-----------------------------
# Helper: BED (0-based start, end exclusive or inclusive doesn't matter much if consistent)
# We convert to 1-based inclusive GRanges: start = start0+1, end = end
#-----------------------------
bed_to_gr <- function(dt, keep_chr = NULL) {
  stopifnot(ncol(dt) >= 3)
  dt <- copy(dt)
  setnames(dt, 1:3, c("chr","start0","end"))
  gr <- GRanges(
    seqnames = dt$chr,
    ranges   = IRanges(start = dt$start0 + 1L, end = dt$end)
  )
  if (!is.null(keep_chr)) gr <- gr[seqnames(gr) %in% keep_chr]
  gr
}

```

# Introduction

To evaluate how **DNase peaks relate to other cCRE peaks**, we performed two overlap analyses: **base-pair overlap** and **peak-level overlap**.

- The DNase peaks: 

file used: `/project/spott/kevinluo/Fiber_seq/data/DNaseseq/GM12878/DNaseseq_GM12878_peaks_ENCFF195QAV.bed.gz`

- cCRE peaks: 

file used: `"/project/spott/kevinluo/Fiber_seq/data/ENCODE_CREs/cCREs_v4/LCLs/GM12878_cCREs_ENCFF428XFI_ENCFF280PUF_ENCFF469WVA_ENCFF644EEX.bed"`

# Results

## GM12878

### 1. Base-pair (bp) overlap

```{r fig.height=6, fig.width=4}

#-----------------------------
# 1) Load DNase peaks
#-----------------------------
DNasepeak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/DNaseseq/GM12878/DNaseseq_GM12878_peaks_ENCFF195QAV.bed.gz",
  showProgress = FALSE
)

DNase_gr_raw <- bed_to_gr(DNasepeak_df, keep_chr)

# Raw counts (peak-based)
n_DNase_raw <- length(DNase_gr_raw)

# Reduced counts (bp-based; avoids double-counting bp)
DNase_gr_red <- reduce(DNase_gr_raw, ignore.strand = TRUE)
n_DNase_red  <- length(DNase_gr_red)


#-----------------------------
# 2) Load Other peaks (possibly multiple .bed.gz files)
#-----------------------------
peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ENCODE_CREs/cCREs_v4/LCLs/GM12878_cCREs_ENCFF428XFI_ENCFF280PUF_ENCFF469WVA_ENCFF644EEX.bed",
  showProgress = FALSE
)
peak_df <- peak_df[peak_df$V10 !="Low-DNase",]
peak_df <- peak_df[peak_df$V1 %in% paste0("chr",1:22),]

peak_gr_raw <- bed_to_gr(peak_df, keep_chr)

n_other_raw <- length(peak_gr_raw)
peak_gr_red <- reduce(peak_gr_raw, ignore.strand = TRUE)
n_other_red <- length(peak_gr_red)

#-----------------------------
# A) BP overlap (on reduced ranges)
#-----------------------------
ov_red <- intersect(DNase_gr_red, peak_gr_red, ignore.strand = TRUE)

bp_DNase_total  <- sum(width(DNase_gr_red))
bp_other_total <- sum(width(peak_gr_red))
bp_overlap     <- sum(width(ov_red))

cat("\n=== BP overlap (reduced ranges) ===\n")
cat("DNase total bp (reduced):  ", bp_DNase_total,  "\n")
cat("cCRE total bp (reduced): ", bp_other_total, "\n")
cat("Overlap bp:               ", bp_overlap,     "\n")
cat("% DNase bp in cCRE:       ", bp_overlap/bp_DNase_total*100,       "\n")
cat("% cCRE bp in DNase:       ", bp_overlap/bp_other_total*100,       "\n")

overlap_summary <- data.table(
  metric = c("all_cCRE bp in DNase","DNase bp in all_cCRE"),
  percent = c(
    bp_overlap / bp_other_total * 100,
    bp_overlap / bp_DNase_total * 100
  )
)


base_bar_theme <- theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.margin = margin(t = 12, r = 10, b = 5, l = 5)
  )


# Plot percentage barplot
p_bp <- ggplot(overlap_summary, aes(x = metric, y = percent, fill = metric)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = sprintf("%.1f%%", percent)),
    vjust = -0.3, size = 4.5
  ) +
  labs(
    x = NULL,
    y = "Percentage (%)",
    title = "base-pair"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.12))
  ) +
  coord_cartesian(clip = "off") +
  base_bar_theme

```


```{r fig.height=6, fig.width=12}

#-----------------------------
# 1) Build cCRE GRanges with V10 labels + filter
#-----------------------------
peak_gr <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3),
  V10      = peak_df$V10
)
peak_gr <- peak_gr[
  seqnames(peak_gr) %in% keep_chr &
    !is.na(peak_gr$V10) & peak_gr$V10 != ""
]


#-----------------------------
# 2) Reduce within each V10
#-----------------------------
peak_gr_by_v10 <- split(peak_gr, peak_gr$V10)
peak_gr_red_by_v10 <- endoapply(peak_gr_by_v10, reduce, ignore.strand = TRUE)

#-----------------------------
# 3) Compute bp overlap per V10 (ONE pass)
#-----------------------------
overlap_by_v10 <- rbindlist(lapply(names(peak_gr_red_by_v10), function(v) {
  gr_v <- peak_gr_red_by_v10[[v]]
  
  bp_v_total   <- sum(width(gr_v))
  bp_v_overlap <- sum(width(intersect(DNase_gr_red, gr_v, ignore.strand = TRUE)))
  
  data.table(
    V10 = v,
    bp_v_total = bp_v_total,
    bp_overlap = bp_v_overlap,
    pct_DNase_in_V10 = if (bp_DNase_total > 0) 100 * bp_v_overlap / bp_DNase_total else NA_real_,
    pct_V10_covered_by_DNase = if (bp_v_total > 0) 100 * bp_v_overlap / bp_v_total else NA_real_
  )
}))

# Define a consistent order (e.g., based on pct_DNase_in_V10)
v10_order <- overlap_by_v10[order(-pct_DNase_in_V10), V10]
overlap_by_v10[, V10 := factor(V10, levels = v10_order)]

#-----------------------------
# 4) Plot A: composition of DNase bp across V10
#-----------------------------
pA <- ggplot(overlap_by_v10,
             aes(x = V10, y = pct_DNase_in_V10)) +
  geom_col(width = 0.7) +
  coord_flip() +
  geom_text(aes(label = sprintf("%.1f%%", pct_DNase_in_V10)),
            hjust = -0.05, size = 4) +
  theme_classic(base_size = 14) +
  labs(
    x = "cCRE category",
    y = "% of DNase bp overlapped"
  )

#-----------------------------
# 5) Plot B: coverage of each V10 by DNase
#-----------------------------
pB <- ggplot(overlap_by_v10,
             aes(x = V10, y = pct_V10_covered_by_DNase)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", pct_V10_covered_by_DNase)),
            hjust = -0.05, size = 4) +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    x = "cCRE category",
    y = "% bp of category covered by DNase"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

#-----------------------------
# 6) Combine into one figure (optional)
#-----------------------------
(pA | pB) +
  plot_annotation(title = "DNase vs cCRE overlap by category (bp-based)")

```


### 2. Peak-level overlap -- min overlap 100bp 

```{r fig.height=6, fig.width=12}

min_ov_bp <- 100L  # <-- overlap definition for "peak overlap"

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many DNase peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------

peak_gr_raw <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3),
  V10      = peak_df$V10
)

hitsoverlap <- findOverlaps(
  DNase_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

DNase_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_DNase_with_overlap  <- length(DNase_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)

prop_DNase_overlap  <- n_DNase_with_overlap / n_DNase_raw
prop_other_overlap <- n_other_with_overlap / n_other_raw

cat("\n=== Peak overlap proportions (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat(sprintf("DNase peaks overlapping all_cCRE: %d / %d (%.2f%%)\n",
            n_DNase_with_overlap, n_DNase_raw, 100 * prop_DNase_overlap))
cat(sprintf("all_cCRE peaks overlapping DNase: %d / %d (%.2f%%)\n",
            n_other_with_overlap, n_other_raw, 100 * prop_other_overlap))

overlap_summary <- data.table(
  group = c("DNase peaks overlapping all_cCRE",
            "all_cCRE peaks overlapping DNase"),
  n_overlap = c(n_DNase_with_overlap, n_other_with_overlap),
  n_total   = c(n_DNase_raw, n_other_raw)
)

overlap_summary[, percent := 100 * n_overlap / n_total]

p2 <- ggplot(overlap_summary, aes(x = group, y = percent, fill = group)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = sprintf("%.1f%%\n(%d/%d)", percent, n_overlap, n_total)),
    vjust = -0.3, size = 4.5
  ) +
  labs(
    x = NULL,
    y = "Overlapping peaks (%)",
    title = "Peak-level"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.18))
  ) +
  coord_cartesian(clip = "off") +
  base_bar_theme

p_bp + p2
```

```{r fig.height=6, fig.width=12}

# Keep valid categories only
peak_gr_raw <- peak_gr_raw[!is.na(peak_gr_raw$V10) & peak_gr_raw$V10 != ""]

# Split cCRE peaks by V10 category
peak_by_v10 <- split(peak_gr_raw, peak_gr_raw$V10)

peak_categories <- sort(unique(mcols(peak_gr_raw)$V10))

overlap_summary <- rbindlist(lapply(peak_categories, function(cat) {
  
  gr_v <- peak_gr_raw[mcols(peak_gr_raw)$V10 == cat]
  
  hits <- findOverlaps(
    DNase_gr_raw, gr_v,
    ignore.strand = TRUE,
    minoverlap = min_ov_bp
  )
  
  n_peak_total <- length(gr_v)
  n_DNase_total <- length(DNase_gr_raw)
  
  if (length(hits) == 0) {
    n_peak_overlap <- 0
    n_DNase_overlap <- 0
  } else {
    n_peak_overlap <- length(unique(subjectHits(hits)))  # peaks in this category hit by DNase
    n_DNase_overlap <- length(unique(queryHits(hits)))    # DNases hit by peaks in this category
  }
  
  data.table(
    category = cat,
    n_peak_total = n_peak_total,
    n_peak_overlap = n_peak_overlap,
    prop_peak_overlap = n_peak_overlap / n_peak_total,
    n_DNase_total = n_DNase_total,
    n_DNase_overlap = n_DNase_overlap,
    prop_DNase_overlap = n_DNase_overlap / n_DNase_total
  )
}))

overlap_summary[, `:=`(
  peak_overlap_pct = prop_peak_overlap * 100,
  DNase_overlap_pct = prop_DNase_overlap * 100
)]

plot_dt <- melt(
  overlap_summary[, .(
    category,
    `Peaks overlapping DNase (%)` = peak_overlap_pct,
    `DNase overlapping peaks (%)` = DNase_overlap_pct
  )],
  id.vars = "category",
  variable.name = "metric",
  value.name = "percent"
)

# optional: order categories by one metric
overlap_summary[, category := factor(
  category,
  levels = overlap_summary[order(DNase_overlap_pct, decreasing = T), category]
)]

p1 <- ggplot(
  overlap_summary,
  aes(x = DNase_overlap_pct, y = category)
) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", DNase_overlap_pct)),
            hjust = -0.1, size = 3.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "DNase overlapping peaks (%)",
    y = NULL,
    #title = "Proportion of DNase overlapping each peak category"
  ) +
  theme_classic(base_size = 13)

p2 <- ggplot(
  overlap_summary,
  aes(x = peak_overlap_pct, y = category)
) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", peak_overlap_pct)),
            hjust = -0.1, size = 3.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "Peaks overlapping DNase (%)",
    y = NULL,
    #title = "Proportion of peaks overlapping DNase"
  ) +
  theme_classic(base_size = 13)

(p1 | p2) +
  plot_annotation(title = "DNase vs cCRE overlap by category (peak-level)")




```

### 3. Peak-level overlap -- min overlap 10bp 

```{r fig.height=6, fig.width=12}

min_ov_bp <- 10L  # <-- overlap definition for "peak overlap"

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many DNase peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------

peak_gr_raw <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3),
  V10      = peak_df$V10
)

hitsoverlap <- findOverlaps(
  DNase_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

DNase_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_DNase_with_overlap  <- length(DNase_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)

prop_DNase_overlap  <- n_DNase_with_overlap / n_DNase_raw
prop_other_overlap <- n_other_with_overlap / n_other_raw

cat("\n=== Peak overlap proportions (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat(sprintf("DNase peaks overlapping all_cCRE: %d / %d (%.2f%%)\n",
            n_DNase_with_overlap, n_DNase_raw, 100 * prop_DNase_overlap))
cat(sprintf("all_cCRE peaks overlapping DNase: %d / %d (%.2f%%)\n",
            n_other_with_overlap, n_other_raw, 100 * prop_other_overlap))

overlap_summary <- data.table(
  group = c("DNase peaks overlapping all_cCRE",
            "all_cCRE peaks overlapping DNase"),
  n_overlap = c(n_DNase_with_overlap, n_other_with_overlap),
  n_total   = c(n_DNase_raw, n_other_raw)
)

overlap_summary[, percent := 100 * n_overlap / n_total]

p2 <- ggplot(overlap_summary, aes(x = group, y = percent, fill = group)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = sprintf("%.1f%%\n(%d/%d)", percent, n_overlap, n_total)),
    vjust = -0.3, size = 4.5
  ) +
  labs(
    x = NULL,
    y = "Overlapping peaks (%)",
    title = "Peak-level"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.18))
  ) +
  coord_cartesian(clip = "off") +
  base_bar_theme

p_bp + p2
```

```{r fig.height=6, fig.width=12}

# Keep valid categories only
peak_gr_raw <- peak_gr_raw[!is.na(peak_gr_raw$V10) & peak_gr_raw$V10 != ""]

# Split cCRE peaks by V10 category
peak_by_v10 <- split(peak_gr_raw, peak_gr_raw$V10)

peak_categories <- sort(unique(mcols(peak_gr_raw)$V10))

overlap_summary <- rbindlist(lapply(peak_categories, function(cat) {
  
  gr_v <- peak_gr_raw[mcols(peak_gr_raw)$V10 == cat]
  
  hits <- findOverlaps(
    DNase_gr_raw, gr_v,
    ignore.strand = TRUE,
    minoverlap = min_ov_bp
  )
  
  n_peak_total <- length(gr_v)
  n_DNase_total <- length(DNase_gr_raw)
  
  if (length(hits) == 0) {
    n_peak_overlap <- 0
    n_DNase_overlap <- 0
  } else {
    n_peak_overlap <- length(unique(subjectHits(hits)))  # peaks in this category hit by DNase
    n_DNase_overlap <- length(unique(queryHits(hits)))    # DNases hit by peaks in this category
  }
  
  data.table(
    category = cat,
    n_peak_total = n_peak_total,
    n_peak_overlap = n_peak_overlap,
    prop_peak_overlap = n_peak_overlap / n_peak_total,
    n_DNase_total = n_DNase_total,
    n_DNase_overlap = n_DNase_overlap,
    prop_DNase_overlap = n_DNase_overlap / n_DNase_total
  )
}))

overlap_summary[, `:=`(
  peak_overlap_pct = prop_peak_overlap * 100,
  DNase_overlap_pct = prop_DNase_overlap * 100
)]

plot_dt <- melt(
  overlap_summary[, .(
    category,
    `Peaks overlapping DNase (%)` = peak_overlap_pct,
    `DNase overlapping peaks (%)` = DNase_overlap_pct
  )],
  id.vars = "category",
  variable.name = "metric",
  value.name = "percent"
)

# optional: order categories by one metric
overlap_summary[, category := factor(
  category,
  levels = overlap_summary[order(DNase_overlap_pct, decreasing = T), category]
)]

p1 <- ggplot(
  overlap_summary,
  aes(x = DNase_overlap_pct, y = category)
) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", DNase_overlap_pct)),
            hjust = -0.1, size = 3.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "DNase overlapping peaks (%)",
    y = NULL,
    #title = "Proportion of DNase overlapping each peak category"
  ) +
  theme_classic(base_size = 13)

p2 <- ggplot(
  overlap_summary,
  aes(x = peak_overlap_pct, y = category)
) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", peak_overlap_pct)),
            hjust = -0.1, size = 3.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "Peaks overlapping DNase (%)",
    y = NULL,
    #title = "Proportion of peaks overlapping DNase"
  ) +
  theme_classic(base_size = 13)

(p1 | p2) +
  plot_annotation(title = "DNase vs cCRE overlap by category (peak-level)")




```

## GM12891

### 1. Base-pair (bp) overlap

```{r fig.height=6, fig.width=4}

#-----------------------------
# 1) Load DNase peaks
#-----------------------------
DNasepeak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/DNaseseq/GM12891/DNase_GM12891_peaks_ENCFF087IIE.bed.gz",
  showProgress = FALSE
)

DNase_gr_raw <- bed_to_gr(DNasepeak_df, keep_chr)

# Raw counts (peak-based)
n_DNase_raw <- length(DNase_gr_raw)

# Reduced counts (bp-based; avoids double-counting bp)
DNase_gr_red <- reduce(DNase_gr_raw, ignore.strand = TRUE)
n_DNase_red  <- length(DNase_gr_red)


#-----------------------------
# 2) Load Other peaks (possibly multiple .bed.gz files)
#-----------------------------
peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ENCODE_CREs/cCREs_v4/LCLs/GM12891_cCREs_ENCFF933EUW.bed",
  showProgress = FALSE
)
peak_df <- peak_df[peak_df$V10 !="Low-DNase",]
peak_df <- peak_df[peak_df$V1 %in% paste0("chr",1:22),]

peak_gr_raw <- bed_to_gr(peak_df, keep_chr)

n_other_raw <- length(peak_gr_raw)
peak_gr_red <- reduce(peak_gr_raw, ignore.strand = TRUE)
n_other_red <- length(peak_gr_red)

#-----------------------------
# A) BP overlap (on reduced ranges)
#-----------------------------
ov_red <- intersect(DNase_gr_red, peak_gr_red, ignore.strand = TRUE)

bp_DNase_total  <- sum(width(DNase_gr_red))
bp_other_total <- sum(width(peak_gr_red))
bp_overlap     <- sum(width(ov_red))

cat("\n=== BP overlap (reduced ranges) ===\n")
cat("DNase total bp (reduced):  ", bp_DNase_total,  "\n")
cat("cCRE total bp (reduced): ", bp_other_total, "\n")
cat("Overlap bp:               ", bp_overlap,     "\n")
cat("% DNase bp in cCRE:       ", bp_overlap/bp_DNase_total*100,       "\n")
cat("% cCRE bp in DNase:       ", bp_overlap/bp_other_total*100,       "\n")

overlap_summary <- data.table(
  metric = c("all_cCRE bp in DNase","DNase bp in all_cCRE"),
  percent = c(
    bp_overlap / bp_other_total * 100,
    bp_overlap / bp_DNase_total * 100
  )
)


base_bar_theme <- theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.margin = margin(t = 12, r = 10, b = 5, l = 5)
  )


# Plot percentage barplot
p_bp <- ggplot(overlap_summary, aes(x = metric, y = percent, fill = metric)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = sprintf("%.1f%%", percent)),
    vjust = -0.3, size = 4.5
  ) +
  labs(
    x = NULL,
    y = "Percentage (%)",
    title = "base-pair"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.12))
  ) +
  coord_cartesian(clip = "off") +
  base_bar_theme

```


```{r fig.height=6, fig.width=12}

#-----------------------------
# 1) Build cCRE GRanges with V10 labels + filter
#-----------------------------
peak_gr <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3),
  V10      = peak_df$V10
)
peak_gr <- peak_gr[
  seqnames(peak_gr) %in% keep_chr &
    !is.na(peak_gr$V10) & peak_gr$V10 != ""
]


#-----------------------------
# 2) Reduce within each V10
#-----------------------------
peak_gr_by_v10 <- split(peak_gr, peak_gr$V10)
peak_gr_red_by_v10 <- endoapply(peak_gr_by_v10, reduce, ignore.strand = TRUE)

#-----------------------------
# 3) Compute bp overlap per V10 (ONE pass)
#-----------------------------
overlap_by_v10 <- rbindlist(lapply(names(peak_gr_red_by_v10), function(v) {
  gr_v <- peak_gr_red_by_v10[[v]]
  
  bp_v_total   <- sum(width(gr_v))
  bp_v_overlap <- sum(width(intersect(DNase_gr_red, gr_v, ignore.strand = TRUE)))
  
  data.table(
    V10 = v,
    bp_v_total = bp_v_total,
    bp_overlap = bp_v_overlap,
    pct_DNase_in_V10 = if (bp_DNase_total > 0) 100 * bp_v_overlap / bp_DNase_total else NA_real_,
    pct_V10_covered_by_DNase = if (bp_v_total > 0) 100 * bp_v_overlap / bp_v_total else NA_real_
  )
}))

# Define a consistent order (e.g., based on pct_DNase_in_V10)
v10_order <- overlap_by_v10[order(-pct_DNase_in_V10), V10]
overlap_by_v10[, V10 := factor(V10, levels = v10_order)]

#-----------------------------
# 4) Plot A: composition of DNase bp across V10
#-----------------------------
pA <- ggplot(overlap_by_v10,
             aes(x = V10, y = pct_DNase_in_V10)) +
  geom_col(width = 0.7) +
  coord_flip() +
  geom_text(aes(label = sprintf("%.1f%%", pct_DNase_in_V10)),
            hjust = -0.05, size = 4) +
  theme_classic(base_size = 14) +
  labs(
    x = "cCRE category",
    y = "% of DNase bp overlapped"
  )

#-----------------------------
# 5) Plot B: coverage of each V10 by DNase
#-----------------------------
pB <- ggplot(overlap_by_v10,
             aes(x = V10, y = pct_V10_covered_by_DNase)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", pct_V10_covered_by_DNase)),
            hjust = -0.05, size = 4) +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    x = "cCRE category",
    y = "% bp of category covered by DNase"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

#-----------------------------
# 6) Combine into one figure (optional)
#-----------------------------
(pA | pB) +
  plot_annotation(title = "DNase vs cCRE overlap by category (bp-based)")

```


### 2. Peak-level overlap -- min overlap 100bp

```{r fig.height=6, fig.width=12}

min_ov_bp <- 100L

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many DNase peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------

peak_gr_raw <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3),
  V10      = peak_df$V10
)

hitsoverlap <- findOverlaps(
  DNase_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

DNase_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_DNase_with_overlap  <- length(DNase_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)

prop_DNase_overlap  <- n_DNase_with_overlap / n_DNase_raw
prop_other_overlap <- n_other_with_overlap / n_other_raw

cat("\n=== Peak overlap proportions (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat(sprintf("DNase peaks overlapping all_cCRE: %d / %d (%.2f%%)\n",
            n_DNase_with_overlap, n_DNase_raw, 100 * prop_DNase_overlap))
cat(sprintf("all_cCRE peaks overlapping DNase: %d / %d (%.2f%%)\n",
            n_other_with_overlap, n_other_raw, 100 * prop_other_overlap))

overlap_summary <- data.table(
  group = c("DNase peaks overlapping all_cCRE",
            "all_cCRE peaks overlapping DNase"),
  n_overlap = c(n_DNase_with_overlap, n_other_with_overlap),
  n_total   = c(n_DNase_raw, n_other_raw)
)

overlap_summary[, percent := 100 * n_overlap / n_total]

p2 <- ggplot(overlap_summary, aes(x = group, y = percent, fill = group)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = sprintf("%.1f%%\n(%d/%d)", percent, n_overlap, n_total)),
    vjust = -0.3, size = 4.5
  ) +
  labs(
    x = NULL,
    y = "Overlapping peaks (%)",
    title = "Peak-level"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.18))
  ) +
  coord_cartesian(clip = "off") +
  base_bar_theme

p_bp + p2
```

```{r fig.height=6, fig.width=12}

# Keep valid categories only
peak_gr_raw <- peak_gr_raw[!is.na(peak_gr_raw$V10) & peak_gr_raw$V10 != ""]

# Split cCRE peaks by V10 category
peak_by_v10 <- split(peak_gr_raw, peak_gr_raw$V10)

peak_categories <- sort(unique(mcols(peak_gr_raw)$V10))

overlap_summary <- rbindlist(lapply(peak_categories, function(cat) {
  
  gr_v <- peak_gr_raw[mcols(peak_gr_raw)$V10 == cat]
  
  hits <- findOverlaps(
    DNase_gr_raw, gr_v,
    ignore.strand = TRUE,
    minoverlap = min_ov_bp
  )
  
  n_peak_total <- length(gr_v)
  n_DNase_total <- length(DNase_gr_raw)
  
  if (length(hits) == 0) {
    n_peak_overlap <- 0
    n_DNase_overlap <- 0
  } else {
    n_peak_overlap <- length(unique(subjectHits(hits)))  # peaks in this category hit by DNase
    n_DNase_overlap <- length(unique(queryHits(hits)))    # DNases hit by peaks in this category
  }
  
  data.table(
    category = cat,
    n_peak_total = n_peak_total,
    n_peak_overlap = n_peak_overlap,
    prop_peak_overlap = n_peak_overlap / n_peak_total,
    n_DNase_total = n_DNase_total,
    n_DNase_overlap = n_DNase_overlap,
    prop_DNase_overlap = n_DNase_overlap / n_DNase_total
  )
}))

overlap_summary[, `:=`(
  peak_overlap_pct = prop_peak_overlap * 100,
  DNase_overlap_pct = prop_DNase_overlap * 100
)]

plot_dt <- melt(
  overlap_summary[, .(
    category,
    `Peaks overlapping DNase (%)` = peak_overlap_pct,
    `DNase overlapping peaks (%)` = DNase_overlap_pct
  )],
  id.vars = "category",
  variable.name = "metric",
  value.name = "percent"
)

# optional: order categories by one metric
overlap_summary[, category := factor(
  category,
  levels = overlap_summary[order(DNase_overlap_pct, decreasing = T), category]
)]

p1 <- ggplot(
  overlap_summary,
  aes(x = DNase_overlap_pct, y = category)
) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", DNase_overlap_pct)),
            hjust = -0.1, size = 3.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "DNase overlapping peaks (%)",
    y = NULL,
    #title = "Proportion of DNase overlapping each peak category"
  ) +
  theme_classic(base_size = 13)

p2 <- ggplot(
  overlap_summary,
  aes(x = peak_overlap_pct, y = category)
) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", peak_overlap_pct)),
            hjust = -0.1, size = 3.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "Peaks overlapping DNase (%)",
    y = NULL,
    #title = "Proportion of peaks overlapping DNase"
  ) +
  theme_classic(base_size = 13)

(p1 | p2) +
  plot_annotation(title = "DNase vs cCRE overlap by category (peak-level)")




```


### 3. Peak-level overlap -- min overlap 10bp

```{r fig.height=6, fig.width=12}

min_ov_bp <- 10L

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many DNase peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------

peak_gr_raw <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3),
  V10      = peak_df$V10
)

hitsoverlap <- findOverlaps(
  DNase_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

DNase_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_DNase_with_overlap  <- length(DNase_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)

prop_DNase_overlap  <- n_DNase_with_overlap / n_DNase_raw
prop_other_overlap <- n_other_with_overlap / n_other_raw

cat("\n=== Peak overlap proportions (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat(sprintf("DNase peaks overlapping all_cCRE: %d / %d (%.2f%%)\n",
            n_DNase_with_overlap, n_DNase_raw, 100 * prop_DNase_overlap))
cat(sprintf("all_cCRE peaks overlapping DNase: %d / %d (%.2f%%)\n",
            n_other_with_overlap, n_other_raw, 100 * prop_other_overlap))

overlap_summary <- data.table(
  group = c("DNase peaks overlapping all_cCRE",
            "all_cCRE peaks overlapping DNase"),
  n_overlap = c(n_DNase_with_overlap, n_other_with_overlap),
  n_total   = c(n_DNase_raw, n_other_raw)
)

overlap_summary[, percent := 100 * n_overlap / n_total]

p2 <- ggplot(overlap_summary, aes(x = group, y = percent, fill = group)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = sprintf("%.1f%%\n(%d/%d)", percent, n_overlap, n_total)),
    vjust = -0.3, size = 4.5
  ) +
  labs(
    x = NULL,
    y = "Overlapping peaks (%)",
    title = "Peak-level"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.18))
  ) +
  coord_cartesian(clip = "off") +
  base_bar_theme

p_bp + p2
```

```{r fig.height=6, fig.width=12}

# Keep valid categories only
peak_gr_raw <- peak_gr_raw[!is.na(peak_gr_raw$V10) & peak_gr_raw$V10 != ""]

# Split cCRE peaks by V10 category
peak_by_v10 <- split(peak_gr_raw, peak_gr_raw$V10)

peak_categories <- sort(unique(mcols(peak_gr_raw)$V10))

overlap_summary <- rbindlist(lapply(peak_categories, function(cat) {
  
  gr_v <- peak_gr_raw[mcols(peak_gr_raw)$V10 == cat]
  
  hits <- findOverlaps(
    DNase_gr_raw, gr_v,
    ignore.strand = TRUE,
    minoverlap = min_ov_bp
  )
  
  n_peak_total <- length(gr_v)
  n_DNase_total <- length(DNase_gr_raw)
  
  if (length(hits) == 0) {
    n_peak_overlap <- 0
    n_DNase_overlap <- 0
  } else {
    n_peak_overlap <- length(unique(subjectHits(hits)))  # peaks in this category hit by DNase
    n_DNase_overlap <- length(unique(queryHits(hits)))    # DNases hit by peaks in this category
  }
  
  data.table(
    category = cat,
    n_peak_total = n_peak_total,
    n_peak_overlap = n_peak_overlap,
    prop_peak_overlap = n_peak_overlap / n_peak_total,
    n_DNase_total = n_DNase_total,
    n_DNase_overlap = n_DNase_overlap,
    prop_DNase_overlap = n_DNase_overlap / n_DNase_total
  )
}))

overlap_summary[, `:=`(
  peak_overlap_pct = prop_peak_overlap * 100,
  DNase_overlap_pct = prop_DNase_overlap * 100
)]

plot_dt <- melt(
  overlap_summary[, .(
    category,
    `Peaks overlapping DNase (%)` = peak_overlap_pct,
    `DNase overlapping peaks (%)` = DNase_overlap_pct
  )],
  id.vars = "category",
  variable.name = "metric",
  value.name = "percent"
)

# optional: order categories by one metric
overlap_summary[, category := factor(
  category,
  levels = overlap_summary[order(DNase_overlap_pct, decreasing = T), category]
)]

p1 <- ggplot(
  overlap_summary,
  aes(x = DNase_overlap_pct, y = category)
) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", DNase_overlap_pct)),
            hjust = -0.1, size = 3.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "DNase overlapping peaks (%)",
    y = NULL,
    #title = "Proportion of DNase overlapping each peak category"
  ) +
  theme_classic(base_size = 13)

p2 <- ggplot(
  overlap_summary,
  aes(x = peak_overlap_pct, y = category)
) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", peak_overlap_pct)),
            hjust = -0.1, size = 3.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "Peaks overlapping DNase (%)",
    y = NULL,
    #title = "Proportion of peaks overlapping DNase"
  ) +
  theme_classic(base_size = 13)

(p1 | p2) +
  plot_annotation(title = "DNase vs cCRE overlap by category (peak-level)")




```
