---
title: "FIRE peak overlap with DNase/ ATAC-seq/ cCREs"
author: "XSun"
date: "2026-03-04"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(GenomicRanges)
  library(IRanges)
  library(VennDiagram)
  library(ggplot2)
  library(patchwork)
  library(gridExtra)
})

keep_chr <- paste0("chr", 1:22)
min_ov_bp <- 100L  # <-- overlap definition for "peak overlap"

#-----------------------------
# Helper: BED (0-based start, end exclusive or inclusive doesn't matter much if consistent)
# We convert to 1-based inclusive GRanges: start = start0+1, end = end
#-----------------------------
bed_to_gr <- function(dt, keep_chr = NULL) {
  stopifnot(ncol(dt) >= 3)
  dt <- copy(dt)
  setnames(dt, 1:3, c("chr","start0","end"))
  gr <- GRanges(
    seqnames = dt$chr,
    ranges   = IRanges(start = dt$start0 + 1L, end = dt$end)
  )
  if (!is.null(keep_chr)) gr <- gr[seqnames(gr) %in% keep_chr]
  gr
}

```

# Introduction

To evaluate how **FIRE peaks relate to other OCR peaks ()**, we performed two overlap analyses: **base-pair overlap** and **peak-level overlap**.

## Data

Data: https://xsun1229.github.io/fiberseq/data.html

For DNase/ ATAC-seq/ cCRE we used GM12878 cell line

## 1. Base-pair (bp) overlap

FIRE peaks and the comparison peak set were first converted to **GRanges objects** using hg38 genomic coordinates (BED format converted from 0-based to 1-based indexing).

To avoid double counting genomic coverage, overlapping peaks within each dataset were **merged using `reduce()`**, producing non-overlapping genomic intervals.

The **shared genomic coverage** between the two datasets was then calculated using:

- `intersect()` to obtain overlapping genomic segments
- `width()` to compute the number of overlapping base pairs

From this analysis we report:

- total base pairs covered by **FIRE peaks**
- total base pairs covered by the **comparison peak set**
- **overlapping base pairs**

This analysis measures **how much genomic space is shared between the two peak sets**.

## 2. Peak-level overlap

To assess overlap at the **individual peak level**, we performed overlap detection using the **original (non-merged) peaks**.

Peak overlaps were identified using `findOverlaps()` with a **minimum overlap threshold of 100 bp**, meaning two peaks were considered overlapping only if they shared **≥100 bp**.

From this analysis we report:

- total number of **FIRE peaks**
- total number of **comparison peaks**
- number of **FIRE peaks overlapping comparison peaks**
- number of **comparison peaks overlapping FIRE peaks**
- total number of **overlapping peak pairs**

Peak-level overlap was summarized using a **two-set Venn diagram**, showing:

- FIRE peaks only
- comparison peaks only
- overlapping peaks

Because genomic intervals can overlap in a **many-to-many manner**, the Venn intersection was defined as **the number of FIRE peaks that overlap the comparison peak set by ≥100 bp**.

# Results

## Data overview -- peak length distribution

### FIRE peaks

```{r fig.height=4, fig.width=8}
#-----------------------------
# 1) Load FIRE peaks
#-----------------------------
FIREpeak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/results/QTL/fireQTL/consensus_peaks/merged_fire_peaks_31samples.bed.gz",
  showProgress = FALSE
)

FIREpeak_df[, peak_length := V3 - V2]

FIRE_gr_raw <- bed_to_gr(FIREpeak_df, keep_chr)

# Raw counts (peak-based)
n_FIRE_raw <- length(FIRE_gr_raw)

# Reduced counts (bp-based; avoids double-counting bp)
FIRE_gr_red <- reduce(FIRE_gr_raw, ignore.strand = TRUE)
n_FIRE_red  <- length(FIRE_gr_red)

p1 <- ggplot(FIREpeak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "All peaks"
  )

p2 <- ggplot(FIREpeak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  coord_cartesian(xlim = c(0, 2000)) +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "Zooming in to 2000 bp"
  )

(p1 + p2) +
  plot_annotation(
    title = "Peak length distribution of FIRE peaks"
  )
```








### DNase peaks

```{r fig.height=6, fig.width=4}

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/DNaseseq/GM12878/DNaseseq_GM12878_peaks_ENCFF759OLD.bed.gz",
  showProgress = FALSE
)

peak_df[, peak_length := V3 - V2]

p1 <- ggplot(peak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "Peak length distribution of DNase"
  )

print(p1)
```


### ATAC-seq peaks

```{r fig.height=4, fig.width=8}

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ATACseq/GM12878/ATACseq_GM12878_IDR_peaks_ENCFF429WTW.bed.gz",
  showProgress = FALSE
)

peak_df[, peak_length := V3 - V2]

p1 <- ggplot(peak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "IDR"
  )

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ATACseq/GM12878/ATACseq_GM12878_pseudoreplicated_peaks_ENCFF748UZH.bed.gz",
  showProgress = FALSE
)

peak_df[, peak_length := V3 - V2]

p2 <- ggplot(peak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "pseudoreplicated"
  )


(p1 + p2) +
  plot_annotation(
    title = "Peak length distribution of ATAC-seq peaks"
  )
```












### cCRE peaks

```{r fig.height=6, fig.width=4}

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ENCODE_CREs/cCREs_v4/LCLs/GM12878_cCREs_ENCFF428XFI_ENCFF280PUF_ENCFF469WVA_ENCFF644EEX.bed",
  showProgress = FALSE
)

peak_df[, peak_length := V3 - V2]
peak_df <- peak_df[peak_df$V10 != "Low-DNase",]

peak_df[, `:=`(
  cCRE_group = fcase(
    V10 == "PLS",         "PLS",
    V10 == "pELS",        "pELS",
    V10 == "dELS",        "dELS",
    V10 == "CA-only",     "CA-only",
    V10 == "TF",          "TF",
    V10 == "CA-TF",       "CA-TF",
    V10 == "CA-CTCF",     "CA-CTCF",
    V10 == "CA-H3K4me3",  "CA-H3K4me3",
    default = "Other"
  ),
  cCRE_desc = fcase(
    V10 == "PLS",         "Promoter-like signature",
    V10 == "pELS",        "Proximal enhancer-like signature",
    V10 == "dELS",        "Distal enhancer-like signature",
    V10 == "CA-only",     "Chromatin accessible only",
    V10 == "TF",          "TF only",
    V10 == "CA-TF",       "Chromatin accessible with TF",
    V10 == "CA-CTCF",     "Chromatin accessible with CTCF",
    V10 == "CA-H3K4me3",  "Chromatin accessible with H3K4me3",
    default = "Other / unknown"
  )
)]

# -----------------------------
# Summary table (by detailed category)
# -----------------------------
summary_table <- peak_df[, .(
  n_peaks        = .N,
  mean_width     = round(mean(peak_length), 1),
  median_width   = round(median(peak_length), 1),
  total_bp       = sum(peak_length)
), by = .(cCRE_group, cCRE_desc)][order(-n_peaks)]



DT::datatable(summary_table,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','cCRE overview '),options = list(pageLength = 10) )

p1 <- ggplot(peak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "Peak length distribution of cCRE peaks"
  )


print(p1)
```


```{r fig.height=8, fig.width=12}

ggplot(peak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  facet_wrap(~V10) +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "Peak length distribution across cCRE categories"
  )


```



## Overlaps

### With DNase -- GM12878

#### 1. Base-pair (bp) overlap

```{r fig.height=6, fig.width=4}

#-----------------------------
# 2) Load Other peaks (possibly multiple .bed.gz files)
#-----------------------------
peak_folder <- "/project/spott/kevinluo/Fiber_seq/data/DNaseseq/GM12878/"

peak_files <- list.files(
  path = peak_folder,
  pattern = "_peaks_.*\\.bed\\.gz$",
  full.names = TRUE
)
stopifnot(length(peak_files) >= 1)

other_list <- lapply(peak_files, fread, showProgress = FALSE)
peak_df <- rbindlist(other_list, use.names = FALSE, fill = TRUE)

peak_gr_raw <- bed_to_gr(peak_df, keep_chr)

n_other_raw <- length(peak_gr_raw)
peak_gr_red <- reduce(peak_gr_raw, ignore.strand = TRUE)
n_other_red <- length(peak_gr_red)

#-----------------------------
# A) BP overlap (on reduced ranges)
#-----------------------------
ov_red <- intersect(FIRE_gr_red, peak_gr_red, ignore.strand = TRUE)

bp_FIRE_total  <- sum(width(FIRE_gr_red))
bp_other_total <- sum(width(peak_gr_red))
bp_overlap     <- sum(width(ov_red))

cat("\n=== BP overlap (reduced ranges) ===\n")
cat("FIRE total bp (reduced):  ", bp_FIRE_total,  "\n")
cat("DNase total bp (reduced): ", bp_other_total, "\n")
cat("Overlap bp:               ", bp_overlap,     "\n")
cat("% FIRE bp in DNase:       ", bp_overlap/bp_FIRE_total*100,       "\n")
cat("% DNase bp in FIRE:       ", bp_overlap/bp_other_total*100,       "\n")

overlap_summary <- data.table(
  metric = c("FIRE bp in DNase", "DNase bp in FIRE"),
  percent = c(
    bp_overlap / bp_FIRE_total * 100,
    bp_overlap / bp_other_total * 100
  )
)

# Plot percentage barplot
ggplot(overlap_summary, aes(x = metric, y = percent, fill = metric)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", percent)),
            vjust = -0.5, size = 5) +
  theme_classic(base_size = 14) +
  labs(
    x = "",
    y = "Percentage (%)",
    #title = "Genomic overlap between FIRE and DNase peaks"
  ) +
  theme(legend.position = "none")
```

#### 2. Peak-level overlap

```{r fig.height=6, fig.width=4}

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many FIRE peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------
hitsoverlap <- findOverlaps(
  FIRE_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

FIRE_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_FIRE_with_overlap  <- length(FIRE_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)


prop_FIRE_overlap  <- n_FIRE_with_overlap / n_FIRE_raw
prop_other_overlap <- n_other_with_overlap / n_other_raw

cat("\n=== Peak overlap proportions (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat(sprintf("FIRE peaks overlapping DNase: %d / %d (%.2f%%)\n",
            n_FIRE_with_overlap, n_FIRE_raw, 100 * prop_FIRE_overlap))
cat(sprintf("DNase peaks overlapping FIRE: %d / %d (%.2f%%)\n",
            n_other_with_overlap, n_other_raw, 100 * prop_other_overlap))

overlap_summary <- data.table(
  group = c("FIRE peaks overlapping DNase",
            "DNase peaks overlapping FIRE"),
  n_overlap = c(n_FIRE_with_overlap, n_other_with_overlap),
  n_total   = c(n_FIRE_raw, n_other_raw)
)

overlap_summary[, percent := 100 * n_overlap / n_total]

ggplot(overlap_summary, aes(x = group, y = percent, fill = group)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = sprintf("%.1f%%\n(%d/%d)", percent, n_overlap, n_total)),
            vjust = -0.3, size = 4.5) +
  theme_classic(base_size = 14) +
  labs(
    x = NULL,
    y = "Overlapping peaks (%)",
    title = paste0("Peak overlap summary (minoverlap = ", min_ov_bp, " bp)")
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  )


```



### With ATAC-seq -- IDR -- GM12878

#### 1. Base-pair (bp) overlap

```{r fig.height=6, fig.width=4}

#-----------------------------
# 2) Load Other peaks (possibly multiple .bed.gz files)
#-----------------------------
peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ATACseq/GM12878/ATACseq_GM12878_IDR_peaks_ENCFF429WTW.bed.gz",
  showProgress = FALSE
)

peak_gr_raw <- bed_to_gr(peak_df, keep_chr)

n_other_raw <- length(peak_gr_raw)
peak_gr_red <- reduce(peak_gr_raw, ignore.strand = TRUE)
n_other_red <- length(peak_gr_red)

#-----------------------------
# A) BP overlap (on reduced ranges)
#-----------------------------
ov_red <- intersect(FIRE_gr_red, peak_gr_red, ignore.strand = TRUE)

bp_FIRE_total  <- sum(width(FIRE_gr_red))
bp_other_total <- sum(width(peak_gr_red))
bp_overlap     <- sum(width(ov_red))

cat("\n=== BP overlap (reduced ranges) ===\n")
cat("FIRE total bp (reduced):  ", bp_FIRE_total,  "\n")
cat("Other total bp (reduced): ", bp_other_total, "\n")
cat("Overlap bp:               ", bp_overlap,     "\n")
cat("% FIRE bp in ATAC-seq:       ", bp_overlap/bp_FIRE_total*100,       "\n")
cat("% ATAC-seq bp in FIRE:       ", bp_overlap/bp_other_total*100,       "\n")

overlap_summary <- data.table(
  metric = c("FIRE bp in ATAC-seq", "ATAC-seq bp in FIRE"),
  percent = c(
    bp_overlap / bp_FIRE_total * 100,
    bp_overlap / bp_other_total * 100
  )
)

# Plot percentage barplot
ggplot(overlap_summary, aes(x = metric, y = percent, fill = metric)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", percent)),
            vjust = -0.5, size = 5) +
  theme_classic(base_size = 14) +
  labs(
    x = "",
    y = "Percentage (%)",
    #title = "Genomic overlap between FIRE and DNase peaks"
  ) +
  theme(legend.position = "none")
```

#### 2. Peak-level overlap

```{r fig.height=6, fig.width=4}

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many FIRE peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------
hitsoverlap <- findOverlaps(
  FIRE_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

FIRE_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_FIRE_with_overlap  <- length(FIRE_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)

prop_FIRE_overlap  <- n_FIRE_with_overlap / n_FIRE_raw
prop_other_overlap <- n_other_with_overlap / n_other_raw

cat("\n=== Peak overlap proportions (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat(sprintf("FIRE peaks overlapping ATAC-seq: %d / %d (%.2f%%)\n",
            n_FIRE_with_overlap, n_FIRE_raw, 100 * prop_FIRE_overlap))
cat(sprintf("ATAC-seq peaks overlapping FIRE: %d / %d (%.2f%%)\n",
            n_other_with_overlap, n_other_raw, 100 * prop_other_overlap))

overlap_summary <- data.table(
  group = c("FIRE peaks overlapping ATAC-seq",
            "ATAC-seq peaks overlapping FIRE"),
  n_overlap = c(n_FIRE_with_overlap, n_other_with_overlap),
  n_total   = c(n_FIRE_raw, n_other_raw)
)

overlap_summary[, percent := 100 * n_overlap / n_total]

ggplot(overlap_summary, aes(x = group, y = percent, fill = group)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = sprintf("%.1f%%\n(%d/%d)", percent, n_overlap, n_total)),
            vjust = -0.3, size = 4.5) +
  theme_classic(base_size = 14) +
  labs(
    x = NULL,
    y = "Overlapping peaks (%)",
    title = paste0("Peak overlap summary (minoverlap = ", min_ov_bp, " bp)")
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  )


```

### With ATAC-seq -- pseudoreplicated -- GM12878

#### 1. Base-pair (bp) overlap

```{r fig.height=6, fig.width=4}

#-----------------------------
# 2) Load Other peaks (possibly multiple .bed.gz files)
#-----------------------------
peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ATACseq/GM12878/ATACseq_GM12878_pseudoreplicated_peaks_ENCFF748UZH.bed.gz",
  showProgress = FALSE
)

peak_gr_raw <- bed_to_gr(peak_df, keep_chr)

n_other_raw <- length(peak_gr_raw)
peak_gr_red <- reduce(peak_gr_raw, ignore.strand = TRUE)
n_other_red <- length(peak_gr_red)

#-----------------------------
# A) BP overlap (on reduced ranges)
#-----------------------------
ov_red <- intersect(FIRE_gr_red, peak_gr_red, ignore.strand = TRUE)

bp_FIRE_total  <- sum(width(FIRE_gr_red))
bp_other_total <- sum(width(peak_gr_red))
bp_overlap     <- sum(width(ov_red))

cat("\n=== BP overlap (reduced ranges) ===\n")
cat("FIRE total bp (reduced):  ", bp_FIRE_total,  "\n")
cat("Other total bp (reduced): ", bp_other_total, "\n")
cat("Overlap bp:               ", bp_overlap,     "\n")
cat("% FIRE bp in ATAC-seq:       ", bp_overlap/bp_FIRE_total*100,       "\n")
cat("% ATAC-seq bp in FIRE:       ", bp_overlap/bp_other_total*100,       "\n")

overlap_summary <- data.table(
  metric = c("FIRE bp in ATAC-seq", "ATAC-seq bp in FIRE"),
  percent = c(
    bp_overlap / bp_FIRE_total * 100,
    bp_overlap / bp_other_total * 100
  )
)

# Plot percentage barplot
ggplot(overlap_summary, aes(x = metric, y = percent, fill = metric)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", percent)),
            vjust = -0.5, size = 5) +
  theme_classic(base_size = 14) +
  labs(
    x = "",
    y = "Percentage (%)",
    #title = "Genomic overlap between FIRE and DNase peaks"
  ) +
  theme(legend.position = "none")
```

#### 2. Peak-level overlap

```{r fig.height=6, fig.width=4}

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many FIRE peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------
hitsoverlap <- findOverlaps(
  FIRE_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

FIRE_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_FIRE_with_overlap  <- length(FIRE_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)

prop_FIRE_overlap  <- n_FIRE_with_overlap / n_FIRE_raw
prop_other_overlap <- n_other_with_overlap / n_other_raw

cat("\n=== Peak overlap proportions (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat(sprintf("FIRE peaks overlapping ATAC-seq: %d / %d (%.2f%%)\n",
            n_FIRE_with_overlap, n_FIRE_raw, 100 * prop_FIRE_overlap))
cat(sprintf("ATAC-seq peaks overlapping FIRE: %d / %d (%.2f%%)\n",
            n_other_with_overlap, n_other_raw, 100 * prop_other_overlap))

overlap_summary <- data.table(
  group = c("FIRE peaks overlapping ATAC-seq",
            "ATAC-seq peaks overlapping FIRE"),
  n_overlap = c(n_FIRE_with_overlap, n_other_with_overlap),
  n_total   = c(n_FIRE_raw, n_other_raw)
)

overlap_summary[, percent := 100 * n_overlap / n_total]

ggplot(overlap_summary, aes(x = group, y = percent, fill = group)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = sprintf("%.1f%%\n(%d/%d)", percent, n_overlap, n_total)),
            vjust = -0.3, size = 4.5) +
  theme_classic(base_size = 14) +
  labs(
    x = NULL,
    y = "Overlapping peaks (%)",
    title = paste0("Peak overlap summary (minoverlap = ", min_ov_bp, " bp)")
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  )


```




### With cCRE -- GM12878

#### 1. Base-pair (bp) overlap

```{r fig.height=6, fig.width=4}

#-----------------------------
# 2) Load Other peaks (possibly multiple .bed.gz files)
#-----------------------------
peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ENCODE_CREs/cCREs_v4/LCLs/GM12878_cCREs_ENCFF428XFI_ENCFF280PUF_ENCFF469WVA_ENCFF644EEX.bed",
  showProgress = FALSE
)
peak_df <- peak_df[peak_df$V10 !="Low-DNase",]
peak_df <- peak_df[peak_df$V1 %in% paste0("chr",1:22),]

peak_gr_raw <- bed_to_gr(peak_df, keep_chr)

n_other_raw <- length(peak_gr_raw)
peak_gr_red <- reduce(peak_gr_raw, ignore.strand = TRUE)
n_other_red <- length(peak_gr_red)

#-----------------------------
# A) BP overlap (on reduced ranges)
#-----------------------------
ov_red <- intersect(FIRE_gr_red, peak_gr_red, ignore.strand = TRUE)

bp_FIRE_total  <- sum(width(FIRE_gr_red))
bp_other_total <- sum(width(peak_gr_red))
bp_overlap     <- sum(width(ov_red))

cat("\n=== BP overlap (reduced ranges) ===\n")
cat("FIRE total bp (reduced):  ", bp_FIRE_total,  "\n")
cat("cCRE total bp (reduced): ", bp_other_total, "\n")
cat("Overlap bp:               ", bp_overlap,     "\n")
cat("% FIRE bp in cCRE:       ", bp_overlap/bp_FIRE_total*100,       "\n")
cat("% cCRE bp in FIRE:       ", bp_overlap/bp_other_total*100,       "\n")

overlap_summary <- data.table(
  metric = c("cCRE bp in FIRE","FIRE bp in cCRE"),
  percent = c(
    bp_overlap / bp_other_total * 100,
    bp_overlap / bp_FIRE_total * 100
  )
)

overlap_summary[, metric := factor(metric,
                                   levels = c("cCRE bp in FIRE",
                                              "FIRE bp in cCRE"))]

# Plot percentage barplot
ggplot(overlap_summary, aes(x = metric, y = percent, fill = metric)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", percent)),
            vjust = -0.5, size = 5) +
  theme_classic(base_size = 14) +
  labs(
    x = "",
    y = "Percentage (%)",
    title = "all cCRE peaks"
  ) +
  theme(legend.position = "none")
```


```{r fig.height=6, fig.width=12}

#-----------------------------
# 1) Build cCRE GRanges with V10 labels + filter
#-----------------------------
peak_gr <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3),
  V10      = peak_df$V10
)
peak_gr <- peak_gr[
  seqnames(peak_gr) %in% keep_chr &
    !is.na(peak_gr$V10) & peak_gr$V10 != ""
]


#-----------------------------
# 2) Reduce within each V10
#-----------------------------
peak_gr_by_v10 <- split(peak_gr, peak_gr$V10)
peak_gr_red_by_v10 <- endoapply(peak_gr_by_v10, reduce, ignore.strand = TRUE)

#-----------------------------
# 3) Compute bp overlap per V10 (ONE pass)
#-----------------------------
overlap_by_v10 <- rbindlist(lapply(names(peak_gr_red_by_v10), function(v) {
  gr_v <- peak_gr_red_by_v10[[v]]
  
  bp_v_total   <- sum(width(gr_v))
  bp_v_overlap <- sum(width(intersect(FIRE_gr_red, gr_v, ignore.strand = TRUE)))
  
  data.table(
    V10 = v,
    bp_v_total = bp_v_total,
    bp_overlap = bp_v_overlap,
    pct_FIRE_in_V10 = if (bp_FIRE_total > 0) 100 * bp_v_overlap / bp_FIRE_total else NA_real_,
    pct_V10_covered_by_FIRE = if (bp_v_total > 0) 100 * bp_v_overlap / bp_v_total else NA_real_
  )
}))

# Define a consistent order (e.g., based on pct_FIRE_in_V10)
v10_order <- overlap_by_v10[order(-pct_FIRE_in_V10), V10]
overlap_by_v10[, V10 := factor(V10, levels = v10_order)]

#-----------------------------
# 4) Plot A: composition of FIRE bp across V10
#-----------------------------
pA <- ggplot(overlap_by_v10,
             aes(x = V10, y = pct_FIRE_in_V10)) +
  geom_col(width = 0.7) +
  coord_flip() +
  geom_text(aes(label = sprintf("%.1f%%", pct_FIRE_in_V10)),
            hjust = -0.05, size = 4) +
  theme_classic(base_size = 14) +
  labs(
    x = "cCRE category",
    y = "% of FIRE bp overlapped"
  )

#-----------------------------
# 5) Plot B: coverage of each V10 by FIRE
#-----------------------------
pB <- ggplot(overlap_by_v10,
             aes(x = V10, y = pct_V10_covered_by_FIRE)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", pct_V10_covered_by_FIRE)),
            hjust = -0.05, size = 4) +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    x = "cCRE category",
    y = "% bp of category covered by FIRE"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

#-----------------------------
# 6) Combine into one figure (optional)
#-----------------------------
(pA | pB) +
  plot_annotation(title = "FIRE vs cCRE overlap by category (bp-based)")

```


#### 2. Peak-level overlap

```{r fig.height=6, fig.width=4}

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many FIRE peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------

peak_gr_raw <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3),
  V10      = peak_df$V10
)

hitsoverlap <- findOverlaps(
  FIRE_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

FIRE_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_FIRE_with_overlap  <- length(FIRE_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)

prop_FIRE_overlap  <- n_FIRE_with_overlap / n_FIRE_raw
prop_other_overlap <- n_other_with_overlap / n_other_raw

cat("\n=== Peak overlap proportions (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat(sprintf("FIRE peaks overlapping all_cCRE: %d / %d (%.2f%%)\n",
            n_FIRE_with_overlap, n_FIRE_raw, 100 * prop_FIRE_overlap))
cat(sprintf("all_cCRE peaks overlapping FIRE: %d / %d (%.2f%%)\n",
            n_other_with_overlap, n_other_raw, 100 * prop_other_overlap))

overlap_summary <- data.table(
  group = c("FIRE peaks overlapping all_cCRE",
            "all_cCRE peaks overlapping FIRE"),
  n_overlap = c(n_FIRE_with_overlap, n_other_with_overlap),
  n_total   = c(n_FIRE_raw, n_other_raw)
)

overlap_summary[, percent := 100 * n_overlap / n_total]

ggplot(overlap_summary, aes(x = group, y = percent, fill = group)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = sprintf("%.1f%%\n(%d/%d)", percent, n_overlap, n_total)),
            vjust = -0.3, size = 4.5) +
  theme_classic(base_size = 14) +
  labs(
    x = NULL,
    y = "Overlapping peaks (%)",
    title = paste0("Peak overlap summary (minoverlap = ", min_ov_bp, " bp)")
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  )


```

```{r fig.height=6, fig.width=12}

# Keep valid categories only
peak_gr_raw <- peak_gr_raw[!is.na(peak_gr_raw$V10) & peak_gr_raw$V10 != ""]

# Split cCRE peaks by V10 category
peak_by_v10 <- split(peak_gr_raw, peak_gr_raw$V10)

min_ov_bp <- 100   # set your threshold here

peak_categories <- sort(unique(mcols(peak_gr_raw)$V10))

overlap_summary <- rbindlist(lapply(peak_categories, function(cat) {
  
  gr_v <- peak_gr_raw[mcols(peak_gr_raw)$V10 == cat]
  
  hits <- findOverlaps(
    FIRE_gr_raw, gr_v,
    ignore.strand = TRUE,
    minoverlap = min_ov_bp
  )
  
  n_peak_total <- length(gr_v)
  n_fire_total <- length(FIRE_gr_raw)
  
  if (length(hits) == 0) {
    n_peak_overlap <- 0
    n_fire_overlap <- 0
  } else {
    n_peak_overlap <- length(unique(subjectHits(hits)))  # peaks in this category hit by FIRE
    n_fire_overlap <- length(unique(queryHits(hits)))    # FIREs hit by peaks in this category
  }
  
  data.table(
    category = cat,
    n_peak_total = n_peak_total,
    n_peak_overlap = n_peak_overlap,
    prop_peak_overlap = n_peak_overlap / n_peak_total,
    n_fire_total = n_fire_total,
    n_fire_overlap = n_fire_overlap,
    prop_fire_overlap = n_fire_overlap / n_fire_total
  )
}))

overlap_summary[, `:=`(
  peak_overlap_pct = prop_peak_overlap * 100,
  fire_overlap_pct = prop_fire_overlap * 100
)]

plot_dt <- melt(
  overlap_summary[, .(
    category,
    `Peaks overlapping FIRE (%)` = peak_overlap_pct,
    `FIRE overlapping peaks (%)` = fire_overlap_pct
  )],
  id.vars = "category",
  variable.name = "metric",
  value.name = "percent"
)

# optional: order categories by one metric
overlap_summary[, category := factor(
  category,
  levels = overlap_summary[order(fire_overlap_pct, decreasing = T), category]
)]

p1 <- ggplot(
  overlap_summary,
  aes(x = fire_overlap_pct, y = category)
) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", fire_overlap_pct)),
            hjust = -0.1, size = 3.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "FIRE overlapping peaks (%)",
    y = NULL,
    title = "Proportion of FIRE overlapping each peak category"
  ) +
  theme_classic(base_size = 13)

p2 <- ggplot(
  overlap_summary,
  aes(x = peak_overlap_pct, y = category)
) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", peak_overlap_pct)),
            hjust = -0.1, size = 3.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "Peaks overlapping FIRE (%)",
    y = NULL,
    title = "Proportion of peaks overlapping FIRE"
  ) +
  theme_classic(base_size = 13)

(p1 | p2) +
  plot_annotation(title = "FIRE vs cCRE overlap by category (peak-level)")




```
