---
title: "FIRE peak overlap with DNase/ ATAC-seq/ cCREs"
author: "XSun"
date: "2026-03-04"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(GenomicRanges)
  library(IRanges)
  library(VennDiagram)
  library(ggplot2)
  library(patchwork)
  library(gridExtra)
})

keep_chr <- paste0("chr", 1:22)
min_ov_bp <- 100L  # <-- overlap definition for "peak overlap"

#-----------------------------
# Helper: BED (0-based start, end exclusive or inclusive doesn't matter much if consistent)
# We convert to 1-based inclusive GRanges: start = start0+1, end = end
#-----------------------------
bed_to_gr <- function(dt, keep_chr = NULL) {
  stopifnot(ncol(dt) >= 3)
  dt <- copy(dt)
  setnames(dt, 1:3, c("chr","start0","end"))
  gr <- GRanges(
    seqnames = dt$chr,
    ranges   = IRanges(start = dt$start0 + 1L, end = dt$end)
  )
  if (!is.null(keep_chr)) gr <- gr[seqnames(gr) %in% keep_chr]
  gr
}

```

# Introduction

To evaluate how **FIRE peaks relate to other OCR peaks ()**, we performed two overlap analyses: **base-pair overlap** and **peak-level overlap**.

## Data

Data: https://xsun1229.github.io/fiberseq/data.html

For DNase/ ATAC-seq/ cCRE we used GM12878 cell line

## 1. Base-pair (bp) overlap

FIRE peaks and the comparison peak set were first converted to **GRanges objects** using hg38 genomic coordinates (BED format converted from 0-based to 1-based indexing).

To avoid double counting genomic coverage, overlapping peaks within each dataset were **merged using `reduce()`**, producing non-overlapping genomic intervals.

The **shared genomic coverage** between the two datasets was then calculated using:

- `intersect()` to obtain overlapping genomic segments
- `width()` to compute the number of overlapping base pairs

From this analysis we report:

- total base pairs covered by **FIRE peaks**
- total base pairs covered by the **comparison peak set**
- **overlapping base pairs**

This analysis measures **how much genomic space is shared between the two peak sets**.

## 2. Peak-level overlap

To assess overlap at the **individual peak level**, we performed overlap detection using the **original (non-merged) peaks**.

Peak overlaps were identified using `findOverlaps()` with a **minimum overlap threshold of 100 bp**, meaning two peaks were considered overlapping only if they shared **≥100 bp**.

From this analysis we report:

- total number of **FIRE peaks**
- total number of **comparison peaks**
- number of **FIRE peaks overlapping comparison peaks**
- number of **comparison peaks overlapping FIRE peaks**
- total number of **overlapping peak pairs**

Peak-level overlap was summarized using a **two-set Venn diagram**, showing:

- FIRE peaks only
- comparison peaks only
- overlapping peaks

Because genomic intervals can overlap in a **many-to-many manner**, the Venn intersection was defined as **the number of FIRE peaks that overlap the comparison peak set by ≥100 bp**.

# Results

## Data overview -- peak length distribution

### FIRE peaks

```{r fig.height=4, fig.width=8}
#-----------------------------
# 1) Load FIRE peaks
#-----------------------------
FIREpeak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/results/QTL/fireQTL/consensus_peaks/merged_fire_peaks_31samples.bed.gz",
  showProgress = FALSE
)

FIREpeak_df[, peak_length := V3 - V2]

FIRE_gr_raw <- bed_to_gr(FIREpeak_df, keep_chr)

# Raw counts (peak-based)
n_FIRE_raw <- length(FIRE_gr_raw)

# Reduced counts (bp-based; avoids double-counting bp)
FIRE_gr_red <- reduce(FIRE_gr_raw, ignore.strand = TRUE)
n_FIRE_red  <- length(FIRE_gr_red)

p1 <- ggplot(FIREpeak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "All peaks"
  )

p2 <- ggplot(FIREpeak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  coord_cartesian(xlim = c(0, 2000)) +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "Zooming in to 2000 bp"
  )

(p1 + p2) +
  plot_annotation(
    title = "Peak length distribution of FIRE peaks"
  )
```








### DNase peaks

```{r fig.height=4, fig.width=4}

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/DNaseseq/GM12878/DNaseseq_GM12878_peaks_ENCFF759OLD.bed.gz",
  showProgress = FALSE
)

peak_df[, peak_length := V3 - V2]

p1 <- ggplot(peak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "Peak length distribution of DNase"
  )

print(p1)
```


### ATAC-seq peaks

```{r fig.height=4, fig.width=8}

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ATACseq/GM12878/ATACseq_GM12878_IDR_peaks_ENCFF429WTW.bed.gz",
  showProgress = FALSE
)

peak_df[, peak_length := V3 - V2]

p1 <- ggplot(peak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "IDR"
  )

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ATACseq/GM12878/ATACseq_GM12878_pseudoreplicated_peaks_ENCFF748UZH.bed.gz",
  showProgress = FALSE
)

peak_df[, peak_length := V3 - V2]

p2 <- ggplot(peak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "pseudoreplicated"
  )


(p1 + p2) +
  plot_annotation(
    title = "Peak length distribution of ATAC-seq peaks"
  )
```












### cCRE peaks

```{r fig.height=4, fig.width=4}

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ENCODE_CREs/cCREs_v4/LCLs/GM12878_cCREs_ENCFF428XFI_ENCFF280PUF_ENCFF469WVA_ENCFF644EEX.bed",
  showProgress = FALSE
)

peak_df[, peak_length := V3 - V2]
peak_df <- peak_df[peak_df$V10 != "Low-DNase",]

peak_df[, `:=`(
  cCRE_group = fcase(
    V10 == "PLS",         "PLS",
    V10 == "pELS",        "pELS",
    V10 == "dELS",        "dELS",
    V10 == "CA-only",     "CA-only",
    V10 == "TF",          "TF",
    V10 == "CA-TF",       "CA-TF",
    V10 == "CA-CTCF",     "CA-CTCF",
    V10 == "CA-H3K4me3",  "CA-H3K4me3",
    default = "Other"
  ),
  cCRE_desc = fcase(
    V10 == "PLS",         "Promoter-like signature",
    V10 == "pELS",        "Proximal enhancer-like signature",
    V10 == "dELS",        "Distal enhancer-like signature",
    V10 == "CA-only",     "Chromatin accessible only",
    V10 == "TF",          "TF only",
    V10 == "CA-TF",       "Chromatin accessible with TF",
    V10 == "CA-CTCF",     "Chromatin accessible with CTCF",
    V10 == "CA-H3K4me3",  "Chromatin accessible with H3K4me3",
    default = "Other / unknown"
  )
)]

# -----------------------------
# Summary table (by detailed category)
# -----------------------------
summary_table <- peak_df[, .(
  n_peaks        = .N,
  mean_width     = round(mean(peak_length), 1),
  median_width   = round(median(peak_length), 1),
  total_bp       = sum(peak_length)
), by = .(cCRE_group, cCRE_desc)][order(-n_peaks)]



DT::datatable(summary_table,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','cCRE overview '),options = list(pageLength = 10) )

p1 <- ggplot(peak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "Peak length distribution of cCRE peaks"
  )


print(p1)
```


```{r fig.height=8, fig.width=12}

ggplot(peak_df, aes(x = peak_length)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black") +
  facet_wrap(~V10) +
  theme_classic(base_size = 14) +
  labs(
    x = "Peak length (bp)",
    y = "Number of peaks",
    title = "Peak length distribution across cCRE categories"
  )


```



## Overlaps

### With DNase -- GM12878

#### 1. Base-pair (bp) overlap

```{r fig.height=4, fig.width=4}

#-----------------------------
# 2) Load Other peaks (possibly multiple .bed.gz files)
#-----------------------------
peak_folder <- "/project/spott/kevinluo/Fiber_seq/data/DNaseseq/GM12878/"

peak_files <- list.files(
  path = peak_folder,
  pattern = "_peaks_.*\\.bed\\.gz$",
  full.names = TRUE
)
stopifnot(length(peak_files) >= 1)

other_list <- lapply(peak_files, fread, showProgress = FALSE)
peak_df <- rbindlist(other_list, use.names = FALSE, fill = TRUE)

peak_gr_raw <- bed_to_gr(peak_df, keep_chr)

n_other_raw <- length(peak_gr_raw)
peak_gr_red <- reduce(peak_gr_raw, ignore.strand = TRUE)
n_other_red <- length(peak_gr_red)

#-----------------------------
# A) BP overlap (on reduced ranges)
#-----------------------------
ov_red <- intersect(FIRE_gr_red, peak_gr_red, ignore.strand = TRUE)

bp_FIRE_total  <- sum(width(FIRE_gr_red))
bp_other_total <- sum(width(peak_gr_red))
bp_overlap     <- sum(width(ov_red))

cat("\n=== BP overlap (reduced ranges) ===\n")
cat("FIRE total bp (reduced):  ", bp_FIRE_total,  "\n")
cat("DNase total bp (reduced): ", bp_other_total, "\n")
cat("Overlap bp:               ", bp_overlap,     "\n")
cat("% FIRE bp in DNase:       ", bp_overlap/bp_FIRE_total*100,       "\n")
cat("% DNase bp in FIRE:       ", bp_overlap/bp_other_total*100,       "\n")

overlap_summary <- data.table(
  metric = c("FIRE bp in DNase", "DNase bp in FIRE"),
  percent = c(
    bp_overlap / bp_FIRE_total * 100,
    bp_overlap / bp_other_total * 100
  )
)

# Plot percentage barplot
ggplot(overlap_summary, aes(x = metric, y = percent, fill = metric)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", percent)),
            vjust = -0.5, size = 5) +
  theme_classic(base_size = 14) +
  labs(
    x = "",
    y = "Percentage (%)",
    #title = "Genomic overlap between FIRE and DNase peaks"
  ) +
  theme(legend.position = "none")
```

#### 2. Peak-level overlap

```{r fig.height=4, fig.width=4}

peak_folder <- "/project/spott/kevinluo/Fiber_seq/data/DNaseseq/GM12878/"

peak_files <- list.files(
  path = peak_folder,
  pattern = "_peaks_.*\\.bed\\.gz$",
  full.names = TRUE
)
stopifnot(length(peak_files) >= 1)

other_list <- lapply(peak_files, fread, showProgress = FALSE)
peak_df <- rbindlist(other_list, use.names = FALSE, fill = TRUE)

gr_peaks <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3)
)


min_bp <- 100L

## --- 1) One overlap for ALL states (fast) ---
hits <- findOverlaps(gr_peaks, FIRE_gr_raw, ignore.strand = TRUE)

# overlap width for each hit
ov_bp <- width(pintersect(gr_peaks[queryHits(hits)], FIRE_gr_raw[subjectHits(hits)]))

# keep only >= min_bp
keep <- ov_bp >= min_bp
qh <- queryHits(hits)[keep]      # peak indices
sh <- subjectHits(hits)[keep]    # chromhmm segment indices

## --- 2) Summarize per state (unique peak hits, unique segment hits) ---
dt_hits <- data.table(
  peak_i = qh,
  seg_i  = sh
)

sum_by_state <- dt_hits[, .(
  OCRpeaks_overlap        = uniqueN(peak_i),
  chromHMMstate_overlap   = uniqueN(seg_i)
), by = state]

## --- 3) Add totals for each state and compute proportions ---
# totals
n_OCRpeaks <- length(gr_peaks)

dt_state_totals <- data.table(
  state = names(table(mcols(gr_hmm)$state)),
  n_chromHMMstate_peaks = as.integer(table(mcols(gr_hmm)$state))
)

summary_table <- merge(dt_state_totals, sum_by_state, by = "state", all.x = TRUE)

# states with no overlap -> NA to 0
for (col in c("OCRpeaks_overlap", "chromHMMstate_overlap")) {
  summary_table[is.na(get(col)), (col) := 0L]
}

summary_table[, `:=`(
  n_OCRpeaks = n_OCRpeaks,
  OCRpeaks_overlap_prop = OCRpeaks_overlap / n_OCRpeaks,
  chromHMMstate_overlap_prop = chromHMMstate_overlap / n_chromHMMstate_peaks
)]

## --- 4) Add annotation + select final columns ---
setDT(chromhmm_annot)

final_table <- merge(
  chromhmm_annot[, .(state, state_id, mnemonic, description)],
  summary_table[, .(
    state,
    n_OCRpeaks,
    n_chromHMMstate_peaks,
    OCRpeaks_overlap,
    OCRpeaks_overlap_prop,
    chromHMMstate_overlap,
    chromHMMstate_overlap_prop
  )],
  by = "state",
  all.x = TRUE
)

# if some states missing in summary_table (shouldn't), set to 0
for (col in c("n_OCRpeaks","n_chromHMMstate_peaks","OCRpeaks_overlap","chromHMMstate_overlap",
              "OCRpeaks_overlap_prop","chromHMMstate_overlap_prop")) {
  final_table[is.na(get(col)), (col) := if (grepl("prop$", col)) 0 else 0L]
}

final_table <- final_table[, .(
  state_id, mnemonic, description,
  n_OCRpeaks,
  n_chromHMMstate_peaks,
  OCRpeaks_overlap,
  OCRpeaks_overlap_prop,
  chromHMMstate_overlap,
  chromHMMstate_overlap_prop
)]

setorder(final_table, state_id)

DT::datatable(final_table,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Overlap summary'),options = list(pageLength = 10) )

final_table[, OCRpeaks_overlap_pct := OCRpeaks_overlap_prop * 100]
final_table[, chromHMMstate_overlap_pct := chromHMMstate_overlap_prop * 100]



final_table[, state_label := paste(state_id, mnemonic, sep = " ")]
final_table[, state_label := factor(state_label,
                                    levels = final_table[order(state_id), state_label])]

final_table <- merge(
  final_table,
  chromhmm_annot[, .(state_id, hex)],
  by = "state_id",
  all.x = TRUE
)

p1 <- ggplot(final_table,
       aes(x = state_label, y = OCRpeaks_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(
    x = "ChromHMM state",
    y = "OCR peaks overlapping (%)",
    title = "Percentage of OCR peaks \noverlap with each ChromHMM state"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


p2 <- ggplot(final_table, aes(x = state_label, y = chromHMMstate_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(x = "ChromHMM state", y = "ChromHMM segments overlapped (%)",
       title = "Percentage of ChromHMM peaks \noverlap with OCR peaks") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p1)
print(p2)
```



### With ATAC-seq -- IDR -- GM12878

#### 1. Base-pair (bp) overlap

```{r fig.height=4, fig.width=4}

#-----------------------------
# 2) Load Other peaks (possibly multiple .bed.gz files)
#-----------------------------
peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ATACseq/GM12878/ATACseq_GM12878_IDR_peaks_ENCFF429WTW.bed.gz",
  showProgress = FALSE
)

peak_gr_raw <- bed_to_gr(peak_df, keep_chr)

n_other_raw <- length(peak_gr_raw)
peak_gr_red <- reduce(peak_gr_raw, ignore.strand = TRUE)
n_other_red <- length(peak_gr_red)

#-----------------------------
# A) BP overlap (on reduced ranges)
#-----------------------------
ov_red <- intersect(FIRE_gr_red, peak_gr_red, ignore.strand = TRUE)

bp_FIRE_total  <- sum(width(FIRE_gr_red))
bp_other_total <- sum(width(peak_gr_red))
bp_overlap     <- sum(width(ov_red))

cat("\n=== BP overlap (reduced ranges) ===\n")
cat("FIRE total bp (reduced):  ", bp_FIRE_total,  "\n")
cat("Other total bp (reduced): ", bp_other_total, "\n")
cat("Overlap bp:               ", bp_overlap,     "\n")
cat("% FIRE bp in ATAC-seq:       ", bp_overlap/bp_FIRE_total*100,       "\n")
cat("% ATAC-seq bp in FIRE:       ", bp_overlap/bp_other_total*100,       "\n")

overlap_summary <- data.table(
  metric = c("FIRE bp in ATAC-seq", "ATAC-seq bp in FIRE"),
  percent = c(
    bp_overlap / bp_FIRE_total * 100,
    bp_overlap / bp_other_total * 100
  )
)

# Plot percentage barplot
ggplot(overlap_summary, aes(x = metric, y = percent, fill = metric)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", percent)),
            vjust = -0.5, size = 5) +
  theme_classic(base_size = 14) +
  labs(
    x = "",
    y = "Percentage (%)",
    #title = "Genomic overlap between FIRE and DNase peaks"
  ) +
  theme(legend.position = "none")
```

#### 2. Peak-level overlap

```{r fig.height=4, fig.width=4}

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many FIRE peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------
hitsoverlap <- findOverlaps(
  FIRE_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

FIRE_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_FIRE_with_overlap  <- length(FIRE_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)

cat("\n=== Peak overlap (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat("FIRE peaks (raw):                    ", n_FIRE_raw, "\n")
cat("ATAC-seq peaks (raw):                   ", n_other_raw, "\n")
cat("Number of overlapping peak pairs:    ", n_pairs_overlap, "\n")


#-----------------------------
# C) 2-set Venn (peak overlap)
# NOTE: With genomic intervals, overlap is not 1-to-1, so "intersection size" is asymmetric.
# Here we define intersection = number of FIRE peaks that overlap Other by >=min_ov_bp.
# We also print the "Other peaks that overlap FIRE" separately above.
#-----------------------------
venn_a_only <- n_FIRE_raw - n_FIRE_with_overlap
venn_b_only <- n_other_raw - n_other_with_overlap
venn_inter  <- n_FIRE_with_overlap  # definition for the Venn plot below

grid::grid.newpage()
VennDiagram::draw.pairwise.venn(
  area1 = n_FIRE_raw,
  area2 = n_other_raw,
  cross.area = venn_inter,
  category = c("FIRE", "ATAC-seq - IDR"),
  scaled = T
)

```

### With ATAC-seq -- pseudoreplicated -- GM12878

#### 1. Base-pair (bp) overlap

```{r fig.height=4, fig.width=4}

#-----------------------------
# 2) Load Other peaks (possibly multiple .bed.gz files)
#-----------------------------
peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ATACseq/GM12878/ATACseq_GM12878_pseudoreplicated_peaks_ENCFF748UZH.bed.gz",
  showProgress = FALSE
)

peak_gr_raw <- bed_to_gr(peak_df, keep_chr)

n_other_raw <- length(peak_gr_raw)
peak_gr_red <- reduce(peak_gr_raw, ignore.strand = TRUE)
n_other_red <- length(peak_gr_red)

#-----------------------------
# A) BP overlap (on reduced ranges)
#-----------------------------
ov_red <- intersect(FIRE_gr_red, peak_gr_red, ignore.strand = TRUE)

bp_FIRE_total  <- sum(width(FIRE_gr_red))
bp_other_total <- sum(width(peak_gr_red))
bp_overlap     <- sum(width(ov_red))

cat("\n=== BP overlap (reduced ranges) ===\n")
cat("FIRE total bp (reduced):  ", bp_FIRE_total,  "\n")
cat("Other total bp (reduced): ", bp_other_total, "\n")
cat("Overlap bp:               ", bp_overlap,     "\n")
cat("% FIRE bp in ATAC-seq:       ", bp_overlap/bp_FIRE_total*100,       "\n")
cat("% ATAC-seq bp in FIRE:       ", bp_overlap/bp_other_total*100,       "\n")

overlap_summary <- data.table(
  metric = c("FIRE bp in ATAC-seq", "ATAC-seq bp in FIRE"),
  percent = c(
    bp_overlap / bp_FIRE_total * 100,
    bp_overlap / bp_other_total * 100
  )
)

# Plot percentage barplot
ggplot(overlap_summary, aes(x = metric, y = percent, fill = metric)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", percent)),
            vjust = -0.5, size = 5) +
  theme_classic(base_size = 14) +
  labs(
    x = "",
    y = "Percentage (%)",
    #title = "Genomic overlap between FIRE and DNase peaks"
  ) +
  theme(legend.position = "none")
```

#### 2. Peak-level overlap

```{r fig.height=4, fig.width=4}

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many FIRE peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------
hitsoverlap <- findOverlaps(
  FIRE_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

FIRE_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_FIRE_with_overlap  <- length(FIRE_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)

cat("\n=== Peak overlap (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat("FIRE peaks (raw):                    ", n_FIRE_raw, "\n")
cat("ATAC-seq peaks (raw):                   ", n_other_raw, "\n")
cat("Number of overlapping peak pairs:    ", n_pairs_overlap, "\n")


#-----------------------------
# C) 2-set Venn (peak overlap)
# NOTE: With genomic intervals, overlap is not 1-to-1, so "intersection size" is asymmetric.
# Here we define intersection = number of FIRE peaks that overlap Other by >=min_ov_bp.
# We also print the "Other peaks that overlap FIRE" separately above.
#-----------------------------
venn_a_only <- n_FIRE_raw - n_FIRE_with_overlap
venn_b_only <- n_other_raw - n_other_with_overlap
venn_inter  <- n_FIRE_with_overlap  # definition for the Venn plot below

grid::grid.newpage()
VennDiagram::draw.pairwise.venn(
  area1 = n_FIRE_raw,
  area2 = n_other_raw,
  cross.area = venn_inter,
  category = c("FIRE", "ATAC-seq - pseudoreplicated"),
  scaled = T
)

```




### With cCRE -- GM12878

#### 1. Base-pair (bp) overlap

```{r fig.height=4, fig.width=4}

#-----------------------------
# 2) Load Other peaks (possibly multiple .bed.gz files)
#-----------------------------
peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ENCODE_CREs/cCREs_v4/LCLs/GM12878_cCREs_ENCFF428XFI_ENCFF280PUF_ENCFF469WVA_ENCFF644EEX.bed",
  showProgress = FALSE
)
peak_df <- peak_df[peak_df$V10 !="Low-DNase",]
peak_df <- peak_df[peak_df$V1 %in% paste0("chr",1:22),]

peak_gr_raw <- bed_to_gr(peak_df, keep_chr)

n_other_raw <- length(peak_gr_raw)
peak_gr_red <- reduce(peak_gr_raw, ignore.strand = TRUE)
n_other_red <- length(peak_gr_red)

#-----------------------------
# A) BP overlap (on reduced ranges)
#-----------------------------
ov_red <- intersect(FIRE_gr_red, peak_gr_red, ignore.strand = TRUE)

bp_FIRE_total  <- sum(width(FIRE_gr_red))
bp_other_total <- sum(width(peak_gr_red))
bp_overlap     <- sum(width(ov_red))

cat("\n=== BP overlap (reduced ranges) ===\n")
cat("FIRE total bp (reduced):  ", bp_FIRE_total,  "\n")
cat("cCRE total bp (reduced): ", bp_other_total, "\n")
cat("Overlap bp:               ", bp_overlap,     "\n")
cat("% FIRE bp in cCRE:       ", bp_overlap/bp_FIRE_total*100,       "\n")
cat("% cCRE bp in FIRE:       ", bp_overlap/bp_other_total*100,       "\n")

overlap_summary <- data.table(
  metric = c("cCRE bp in FIRE","FIRE bp in cCRE"),
  percent = c(
    bp_overlap / bp_other_total * 100,
    bp_overlap / bp_FIRE_total * 100
  )
)

overlap_summary[, metric := factor(metric,
                                   levels = c("cCRE bp in FIRE",
                                              "FIRE bp in cCRE"))]

# Plot percentage barplot
ggplot(overlap_summary, aes(x = metric, y = percent, fill = metric)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", percent)),
            vjust = -0.5, size = 5) +
  theme_classic(base_size = 14) +
  labs(
    x = "",
    y = "Percentage (%)",
    title = "all cCRE peaks"
  ) +
  theme(legend.position = "none")
```


```{r fig.height=6, fig.width=12}

#-----------------------------
# 1) Build cCRE GRanges with V10 labels + filter
#-----------------------------
peak_gr <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3),
  V10      = peak_df$V10
)
peak_gr <- peak_gr[
  seqnames(peak_gr) %in% keep_chr &
    !is.na(peak_gr$V10) & peak_gr$V10 != ""
]

#-----------------------------
# 2) Reduce within each V10
#-----------------------------
peak_gr_by_v10 <- split(peak_gr, peak_gr$V10)
peak_gr_red_by_v10 <- endoapply(peak_gr_by_v10, reduce, ignore.strand = TRUE)

#-----------------------------
# 3) Compute bp overlap per V10 (ONE pass)
#-----------------------------
overlap_by_v10 <- rbindlist(lapply(names(peak_gr_red_by_v10), function(v) {
  gr_v <- peak_gr_red_by_v10[[v]]
  
  bp_v_total   <- sum(width(gr_v))
  bp_v_overlap <- sum(width(intersect(FIRE_gr_red, gr_v, ignore.strand = TRUE)))
  
  data.table(
    V10 = v,
    bp_v_total = bp_v_total,
    bp_overlap = bp_v_overlap,
    pct_FIRE_in_V10 = if (bp_FIRE_total > 0) 100 * bp_v_overlap / bp_FIRE_total else NA_real_,
    pct_V10_covered_by_FIRE = if (bp_v_total > 0) 100 * bp_v_overlap / bp_v_total else NA_real_
  )
}))

# Define a consistent order (e.g., based on pct_FIRE_in_V10)
v10_order <- overlap_by_v10[order(-pct_FIRE_in_V10), V10]
overlap_by_v10[, V10 := factor(V10, levels = v10_order)]

#-----------------------------
# 4) Plot A: composition of FIRE bp across V10
#-----------------------------
pA <- ggplot(overlap_by_v10,
             aes(x = V10, y = pct_FIRE_in_V10)) +
  geom_col(width = 0.7) +
  coord_flip() +
  geom_text(aes(label = sprintf("%.1f%%", pct_FIRE_in_V10)),
            hjust = -0.05, size = 4) +
  theme_classic(base_size = 14) +
  labs(
    x = "cCRE category",
    y = "% of FIRE bp overlapped"
  )

#-----------------------------
# 5) Plot B: coverage of each V10 by FIRE
#-----------------------------
pB <- ggplot(overlap_by_v10,
             aes(x = V10, y = pct_V10_covered_by_FIRE)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", pct_V10_covered_by_FIRE)),
            hjust = -0.05, size = 4) +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    x = "cCRE category",
    y = "% bp of category covered by FIRE"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

#-----------------------------
# 6) Combine into one figure (optional)
#-----------------------------
(pA | pB) +
  plot_annotation(title = "FIRE vs cCRE overlap by category (bp-based)")

```


#### 2. Peak-level overlap

```{r fig.height=4, fig.width=4}

#-----------------------------
# B) Peak overlap counts (min overlap = min_ov_bp) on RAW peaks
#    This answers: how many FIRE peaks overlap ANY other peak by >=100bp (and vice versa)
#-----------------------------
hitsoverlap <- findOverlaps(
  FIRE_gr_raw, peak_gr_raw,
  ignore.strand = TRUE,
  minoverlap = min_ov_bp
)

FIRE_hit_idx  <- unique(queryHits(hitsoverlap))
other_hit_idx <- unique(subjectHits(hitsoverlap))

n_FIRE_with_overlap  <- length(FIRE_hit_idx)
n_other_with_overlap <- length(other_hit_idx)
n_pairs_overlap      <- length(hitsoverlap)

cat("\n=== Peak overlap (RAW peaks; minoverlap =", min_ov_bp, "bp) ===\n")
cat("FIRE peaks (raw):                    ", n_FIRE_raw, "\n")
cat("cCRE peaks (raw):                   ", n_other_raw, "\n")
cat("Number of overlapping peak pairs:    ", n_pairs_overlap, "\n")


#-----------------------------
# C) 2-set Venn (peak overlap)
# NOTE: With genomic intervals, overlap is not 1-to-1, so "intersection size" is asymmetric.
# Here we define intersection = number of FIRE peaks that overlap Other by >=min_ov_bp.
# We also print the "Other peaks that overlap FIRE" separately above.
#-----------------------------
venn_a_only <- n_FIRE_raw - n_FIRE_with_overlap
venn_b_only <- n_other_raw - n_other_with_overlap
venn_inter  <- n_FIRE_with_overlap  # definition for the Venn plot below

grid::grid.newpage()
VennDiagram::draw.pairwise.venn(
  area1 = n_FIRE_raw,
  area2 = n_other_raw,
  cross.area = venn_inter,
  category = c("FIRE", "ALL_cCRE"),
  scaled = T
)

```

```{r fig.height=12, fig.width=12}

if (is.null(mcols(peak_gr_raw)$V10)) {
  peak_gr_raw$V10 <- peak_df$V10
}

# Keep valid categories only
peak_gr_raw <- peak_gr_raw[!is.na(peak_gr_raw$V10) & peak_gr_raw$V10 != ""]

# Split cCRE peaks by V10 category
peak_by_v10 <- split(peak_gr_raw, peak_gr_raw$V10)


venn_grobs <- list()

for (v in names(peak_by_v10)) {
  
  gr_v <- peak_by_v10[[v]]
  
  hits <- findOverlaps(
    FIRE_gr_raw, gr_v,
    ignore.strand = TRUE,
    minoverlap = min_ov_bp
  )
  
  n_FIRE_raw <- length(FIRE_gr_raw)
  n_cat_raw  <- length(gr_v)
  
  n_FIRE_with_overlap <- length(unique(queryHits(hits)))
  n_cat_with_overlap  <- length(unique(subjectHits(hits)))
  
  venn_inter <- n_FIRE_with_overlap  # same convention as before
  
  title_grob <- textGrob(
    paste0("Overlap with ", v),
    gp = gpar(fontsize = 12, fontface = "bold")
  )
  
  # draw.pairwise.venn returns a LIST of grobs -> convert to ONE grob via grobTree()
  venn_parts <- draw.pairwise.venn(
    area1 = n_FIRE_raw,
    area2 = n_cat_raw,
    cross.area = venn_inter,
    category = c("FIRE", paste0("cCRE: ", v)),
    scaled = TRUE,
    cex = 0.9,
    cat.cex = 0.9,
    ind = FALSE
  )

  
  venn_grobs[[v]] <- arrangeGrob(
    grobTree(venn_parts),
    top = title_grob)
}

# Show all Venns in one figure
grid.arrange(grobs = venn_grobs, ncol = 3)



```
