---
title: "Checking footprint QC metrics: original MACS footprint peaks in FIRE"
author: "XSun"
date: "2026-03-17"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---


```{r message=FALSE, warning=F}

library(data.table)
library(GenomicRanges)
library(Rsamtools)
library(Biostrings)
library(ggplot2)
library(patchwork)


fp_lengths <- c("10-30bp", "30-60bp", "60-80bp")
#fp_lengths <- c("60-80bp")
folder_fp <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_peaks/fp_inFIRE_1bp/"

fa <- FaFile("/project/xinhe/xsun/camseq/pipeline/reference/genome/GRCh38.release110.fa")
open(fa)

```


# A/T Composition Analysis of Footprints

We analyzed the nucleotide composition of footprint regions across different footprint length categories (10–30 bp, 30–60 bp, 60–80 bp) using the GRCh38 reference genome.

Our footprint files are `/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_peaks/fp_inFIRE_1bp/`

These footprints overlap FIRE peaks (>1 bp)

- 10–30 bp
- 30–60 bp
- 60–80 bp

reference genome: `/project/xinhe/xsun/camseq/pipeline/reference/genome/GRCh38.release110.fa`

Workflow: 

1. We extracted genomic sequences corresponding to each footprint using the reference genome

2. For each footprint, We computed:
- Number of A (adenine) bases
- Number of T (thymine) bases
- Total footprint width

3. Then derived:
- AT fraction = (A + T) / footprint length
- A fraction = A / footprint length
- T fraction = T / footprint length

4. Generated histograms for:
- AT fraction
- A fraction
- T fraction

5. For equally sized upstream and downstream bins, compute the fractions as before.


<!-- ```{r fig.height=4, fig.width=12} -->

<!-- for (fp_length in fp_lengths) { -->

<!--   fp_file <- file.path( -->
<!--     folder_fp, -->
<!--     paste0("combined_all_chrs_", fp_length, "_macs3_q0.1_maxgap1_peaks_in_fire.bed") -->
<!--   ) -->
<!--   ## ----------------------------- -->
<!--   ## read footprint file -->
<!--   ## ----------------------------- -->
<!--   df_fp <- fread(fp_file, header = FALSE, showProgress = FALSE) -->

<!--   setnames(df_fp, c( -->
<!--     "fp_chr","fp_start","fp_end","fp_name","fp_pileup","fp_strand", -->
<!--     "fp_foldenrichment","fp_log10pValue","fp_log10qValue","fp_peak" -->
<!--   )) -->

<!--   setDT(df_fp) -->

<!--   gr_fp <- GRanges( -->
<!--     seqnames = df_fp$fp_chr, -->
<!--     ranges = IRanges(start = df_fp$fp_start, end = df_fp$fp_end) -->
<!--   ) -->

<!--   seqlevels(gr_fp) <- sub("^chr", "", seqlevels(gr_fp)) -->
<!--   seqnames(gr_fp) <- sub("^chr", "", as.character(seqnames(gr_fp))) -->

<!--   seqs_fp <- getSeq(fa, gr_fp) -->

<!--   base_counts <- letterFrequency(seqs_fp, letters = c("A", "T"), as.prob = FALSE) -->

<!--   df_at <- data.table( -->
<!--     A = base_counts[, "A"], -->
<!--     T = base_counts[, "T"], -->
<!--     width = width(gr_fp) -->
<!--   ) -->

<!--   df_at[, AT_fraction := (A + T) / width] -->
<!--   df_at[, A_fraction := A / width] -->
<!--   df_at[, T_fraction := T / width] -->

<!--   p1 <- ggplot(df_at, aes(x = AT_fraction)) + -->
<!--     geom_histogram(bins = 50) + -->
<!--     theme_classic() + -->
<!--     labs( -->
<!--       x = "A/T fraction in footprint", -->
<!--       y = "Number of footprints", -->
<!--       title = fp_length -->
<!--     ) -->

<!--   p2 <- ggplot(df_at, aes(x = A_fraction)) + -->
<!--     geom_histogram(bins = 50) + -->
<!--     theme_classic() + -->
<!--     labs( -->
<!--       x = "A fraction in footprint", -->
<!--       y = "Number of footprints", -->
<!--       title = fp_length -->
<!--     ) -->

<!--   p3 <- ggplot(df_at, aes(x = T_fraction)) + -->
<!--     geom_histogram(bins = 50) + -->
<!--     theme_classic() + -->
<!--     labs( -->
<!--       x = "T fraction in footprint", -->
<!--       y = "Number of footprints", -->
<!--       title = fp_length -->
<!--     ) -->


<!--   print(p1 + p2 + p3) -->

<!-- } -->


<!-- ``` -->



```{r fig.height=12, fig.width=16}


res_all <- list()

for (fp_length in fp_lengths) {
  
  fp_file <- file.path(
    folder_fp,
    paste0("combined_all_chrs_", fp_length, "_macs3_q0.1_maxgap1_peaks_in_fire.bed")
  )
  
  ## -----------------------------
  ## read footprint file
  ## -----------------------------
  df_fp <- fread(fp_file, header = FALSE, showProgress = FALSE)
  
  setnames(df_fp, c(
    "fp_chr","fp_start","fp_end","fp_name","fp_pileup","fp_strand",
    "fp_foldenrichment","fp_log10pValue","fp_log10qValue","fp_peak"
  ))
  
  setDT(df_fp)
  
  ## -----------------------------
  ## make GRanges for footprints
  ## -----------------------------
  gr_fp <- GRanges(
    seqnames = df_fp$fp_chr,
    ranges = IRanges(start = df_fp$fp_start, end = df_fp$fp_end)
  )
  
  ## match chromosome naming to fasta
  seqlevels(gr_fp) <- sub("^chr", "", seqlevels(gr_fp))
  seqnames(gr_fp) <- sub("^chr", "", as.character(seqnames(gr_fp)))
  
  ## -----------------------------
  ## make equally sized upstream/downstream regions
  ## same size as each footprint
  ## -----------------------------
  fp_width <- width(gr_fp)
  
  gr_up <- GRanges(
    seqnames = seqnames(gr_fp),
    ranges = IRanges(
      start = start(gr_fp) - fp_width,
      end   = start(gr_fp) - 1
    )
  )
  
  gr_down <- GRanges(
    seqnames = seqnames(gr_fp),
    ranges = IRanges(
      start = end(gr_fp) + 1,
      end   = end(gr_fp) + fp_width
    )
  )
  
  ## -----------------------------
  ## remove invalid regions
  ## upstream start must be >= 1
  ## -----------------------------
  keep_valid <- start(gr_up) >= 1 & start(gr_fp) >= 1 & start(gr_down) >= 1
  
  gr_fp <- gr_fp[keep_valid]
  gr_up <- gr_up[keep_valid]
  gr_down <- gr_down[keep_valid]
  df_fp <- df_fp[keep_valid]
  
  ## -----------------------------
  ## extract sequences
  ## -----------------------------
  seqs_fp <- getSeq(fa, gr_fp)
  seqs_up <- getSeq(fa, gr_up)
  seqs_down <- getSeq(fa, gr_down)
  
  ## -----------------------------
  ## count A and T
  ## -----------------------------
  counts_fp <- letterFrequency(seqs_fp, letters = c("A", "T"), as.prob = FALSE)
  counts_up <- letterFrequency(seqs_up, letters = c("A", "T"), as.prob = FALSE)
  counts_down <- letterFrequency(seqs_down, letters = c("A", "T"), as.prob = FALSE)
  
  ## -----------------------------
  ## build result table
  ## -----------------------------
  dt_fp <- data.table(
    fp_name = df_fp$fp_name,
    fp_length = fp_length,
    region = "Footprint",
    width = width(gr_fp),
    A = counts_fp[, "A"],
    T = counts_fp[, "T"]
  )
  dt_fp[, A_fraction := A / width]
  dt_fp[, T_fraction := T / width]
  dt_fp[, AT_fraction := (A + T) / width]
  
  dt_up <- data.table(
    fp_name = df_fp$fp_name,
    fp_length = fp_length,
    region = "Upstream",
    width = width(gr_up),
    A = counts_up[, "A"],
    T = counts_up[, "T"]
  )
  dt_up[, A_fraction := A / width]
  dt_up[, T_fraction := T / width]
  dt_up[, AT_fraction := (A + T) / width]
  
  dt_down <- data.table(
    fp_name = df_fp$fp_name,
    fp_length = fp_length,
    region = "Downstream",
    width = width(gr_down),
    A = counts_down[, "A"],
    T = counts_down[, "T"]
  )
  dt_down[, A_fraction := A / width]
  dt_down[, T_fraction := T / width]
  dt_down[, AT_fraction := (A + T) / width]
  
  dt_all <- rbindlist(list(dt_up, dt_fp, dt_down), use.names = TRUE)
  res_all[[fp_length]] <- dt_all
  
  ## -----------------------------
  ## summary table
  ## -----------------------------
  dt_sum <- dt_all[, .(
    n = .N,
    mean_A_fraction = mean(A_fraction),
    median_A_fraction = median(A_fraction),
    mean_T_fraction = mean(T_fraction),
    median_T_fraction = median(T_fraction),
    mean_AT_fraction = mean(AT_fraction),
    median_AT_fraction = median(AT_fraction)
  ), by = .(fp_length, region)]
  
  ## -----------------------------
  ## plots
  ## -----------------------------
  p1 <- ggplot(dt_all, aes(x = region, y = A_fraction, fill = region)) +
    geom_boxplot(outlier.size = 0.3) +
    theme_classic(base_size = 14) +
    labs(
      title = paste0(fp_length, ": A fraction"),
      x = NULL,
      y = "A fraction"
    )
  
  p2 <- ggplot(dt_all, aes(x = region, y = T_fraction, fill = region)) +
    geom_boxplot(outlier.size = 0.3) +
    theme_classic(base_size = 14) +
    labs(
      title = paste0(fp_length, ": T fraction"),
      x = NULL,
      y = "T fraction"
    )
  
  p3 <- ggplot(dt_all, aes(x = region, y = AT_fraction, fill = region)) +
    geom_boxplot(outlier.size = 0.3) +
    theme_classic(base_size = 14) +
    labs(
      title = paste0(fp_length, ": A+T fraction"),
      x = NULL,
      y = "AT fraction"
    )
  
  p4 <- ggplot(dt_all, aes(x = A_fraction, fill = region)) +
    geom_density(alpha = 0.35) +
    theme_classic(base_size = 14) +
    labs(
      title = paste0(fp_length, ": A fraction"),
      x = "A fraction",
      y = "Density"
    )
  
  p5 <- ggplot(dt_all, aes(x = T_fraction, fill = region)) +
    geom_density(alpha = 0.35) +
    theme_classic(base_size = 14) +
    labs(
      title = paste0(fp_length, ": T fraction"),
      x = "T fraction",
      y = "Density"
    )
  
  p6 <- ggplot(dt_all, aes(x = AT_fraction, fill = region)) +
    geom_density(alpha = 0.35) +
    theme_classic(base_size = 14) +
    labs(
      title = paste0(fp_length, ": A+T fraction"),
      x = "AT fraction",
      y = "Density"
    )
  
  p7 <- ggplot(dt_all, aes(x = A_fraction, fill = region)) +
    geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
    theme_classic(base_size = 14) +
    labs(
      title = paste0(fp_length, ": A fraction"),
      x = "A fraction",
      y = "Count"
    )
  
  p8 <- ggplot(dt_all, aes(x = T_fraction, fill = region)) +
    geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
    theme_classic(base_size = 14) +
    labs(
      title = paste0(fp_length, ": T fraction"),
      x = "T fraction",
      y = "Count"
    )
  
  p9 <- ggplot(dt_all, aes(x = AT_fraction, fill = region)) +
    geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
    theme_classic(base_size = 14) +
    labs(
      title = paste0(fp_length, ": A+T fraction"),
      x = "A+T fraction",
      y = "Count"
    )
  
  g <- (p1 | p2 | p3) / (p4 | p5 | p6) / (p7 | p8 | p9)
  
  
  print(g)
  
  
}


close(fa)


```
