---
title: "Checking the OCR coverage for 15 states from ChromHMM -- peak level coverage"
author: "XSun"
date: "2026-03-05"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r}

suppressPackageStartupMessages({
  library(data.table)
  library(GenomicRanges)
  library(ggplot2)
  library(patchwork)
})

source("/project/spott/xsun/fiberseq/2.annotating_footprints/0.functions.R")

chromhmm_annot <- as.data.table(chromhmm_annot)
chromhmm_annot[, state := paste0(state_id, "_", mnemonic)]
state_titles <- chromhmm_annot[, .(state, title = description)]


```

# Introduction

Here we check how the 15 ChromHMM states are covered by the FIRE/ ATAC-seq/ DNase-seq/ cCRE peaks. 

To assess overlap at the **individual peak level**, we performed overlap detection using the **original (non-merged) peaks**.

Peak overlaps were identified using `findOverlaps()` with a **minimum overlap threshold of 100 bp**, meaning two peaks were considered overlapping only if they shared **≥100 bp**.

Peak-level overlap was summarized using a **two-set Venn diagram**,

# Data

Data: https://xsun1229.github.io/fiberseq/data.html

For DNase/ ATAC-seq/ cCRE we used GM12878 cell line

## ChromHMM peak length distribution

```{r fig.height=12, fig.width=20}

ChromHMM_df <- fread("/project/spott/xsun/fiberseq/2.annotating_footprints/data/E116_15_coreMarks_hg38lift_mnemonics.bed.gz", showProgress = FALSE)
ChromHMM_df <- ChromHMM_df[!is.na(V4) & V4 != ""]
setnames(ChromHMM_df, 1:4, c("chr","start0","end","state"))

dt <- as.data.table(ChromHMM_df)
dt[, len_bp := end - start0]
dt[, state_id := as.integer(sub("_.*", "", state))]
dt <- merge(dt, chromhmm_annot, by = "state_id", all.x = TRUE)
dt[, facet_label := paste0(state_id, ": ", description)]
dt[, facet_label := factor(
  facet_label,
  levels = unique(facet_label[order(state_id)])
)]


ggplot(dt, aes(x = len_bp)) +
  geom_histogram(bins = 60) +
  facet_wrap(~ facet_label, scales = "free_y", ncol = 5) +
  scale_x_log10() +
  labs(
    x = "Segment length (bp, log10)",
    y = "Count",
    title = "ChromHMM segment length distribution"
  ) +
  theme_classic(base_size = 13)

```



## Peak length distribution for others 

https://xsun1229.github.io/fiberseq/fire_overlap_OCRs.html#Data_overview_–_peak_length_distribution

# Results

```{r}

min_bp <- 100

gr_hmm <- GRanges(
  seqnames = ChromHMM_df$chr,
  ranges   = IRanges(start = ChromHMM_df$start0 + 1L, end = ChromHMM_df$end),
  state    = ChromHMM_df$state
)

```


## FIRE peak

```{r fig.height=4, fig.width=6}

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/results/QTL/fireQTL/consensus_peaks/merged_fire_peaks_31samples.bed.gz",
  showProgress = FALSE
)

gr_peaks <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3)
)

all_states <- sort(unique(ChromHMM_df$state))

min_bp <- 100L

## --- 1) One overlap for ALL states (fast) ---
hits <- findOverlaps(gr_peaks, gr_hmm, ignore.strand = TRUE)

# overlap width for each hit
ov_bp <- width(pintersect(gr_peaks[queryHits(hits)], gr_hmm[subjectHits(hits)]))

# keep only >= min_bp
keep <- ov_bp >= min_bp
qh <- queryHits(hits)[keep]      # peak indices
sh <- subjectHits(hits)[keep]    # chromhmm segment indices

# state label for each kept hit
st <- mcols(gr_hmm)$state[sh]

## --- 2) Summarize per state (unique peak hits, unique segment hits) ---
dt_hits <- data.table(
  state = st,
  peak_i = qh,
  seg_i  = sh
)

sum_by_state <- dt_hits[, .(
  OCRpeaks_overlap        = uniqueN(peak_i),
  chromHMMstate_overlap   = uniqueN(seg_i)
), by = state]

## --- 3) Add totals for each state and compute proportions ---
# totals
n_OCRpeaks <- length(gr_peaks)

dt_state_totals <- data.table(
  state = names(table(mcols(gr_hmm)$state)),
  n_chromHMMstate_peaks = as.integer(table(mcols(gr_hmm)$state))
)

summary_table <- merge(dt_state_totals, sum_by_state, by = "state", all.x = TRUE)

# states with no overlap -> NA to 0
for (col in c("OCRpeaks_overlap", "chromHMMstate_overlap")) {
  summary_table[is.na(get(col)), (col) := 0L]
}

summary_table[, `:=`(
  n_OCRpeaks = n_OCRpeaks,
  OCRpeaks_overlap_prop = OCRpeaks_overlap / n_OCRpeaks,
  chromHMMstate_overlap_prop = chromHMMstate_overlap / n_chromHMMstate_peaks
)]

## --- 4) Add annotation + select final columns ---
setDT(chromhmm_annot)

final_table <- merge(
  chromhmm_annot[, .(state, state_id, mnemonic, description)],
  summary_table[, .(
    state,
    n_OCRpeaks,
    n_chromHMMstate_peaks,
    OCRpeaks_overlap,
    OCRpeaks_overlap_prop,
    chromHMMstate_overlap,
    chromHMMstate_overlap_prop
  )],
  by = "state",
  all.x = TRUE
)

# if some states missing in summary_table (shouldn't), set to 0
for (col in c("n_OCRpeaks","n_chromHMMstate_peaks","OCRpeaks_overlap","chromHMMstate_overlap",
              "OCRpeaks_overlap_prop","chromHMMstate_overlap_prop")) {
  final_table[is.na(get(col)), (col) := if (grepl("prop$", col)) 0 else 0L]
}

final_table <- final_table[, .(
  state_id, mnemonic, description,
  n_OCRpeaks,
  n_chromHMMstate_peaks,
  OCRpeaks_overlap,
  OCRpeaks_overlap_prop,
  chromHMMstate_overlap,
  chromHMMstate_overlap_prop
)]

setorder(final_table, state_id)

DT::datatable(final_table,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Overlap summary'),options = list(pageLength = 10) )

final_table[, OCRpeaks_overlap_pct := OCRpeaks_overlap_prop * 100]
final_table[, chromHMMstate_overlap_pct := chromHMMstate_overlap_prop * 100]



final_table[, state_label := paste(state_id, mnemonic, sep = " ")]
final_table[, state_label := factor(state_label,
                                    levels = final_table[order(state_id), state_label])]

final_table <- merge(
  final_table,
  chromhmm_annot[, .(state_id, hex)],
  by = "state_id",
  all.x = TRUE
)

p1 <- ggplot(final_table,
       aes(x = state_label, y = OCRpeaks_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(
    x = "ChromHMM state",
    y = "OCR peaks overlapping (%)",
    title = "Percentage of OCR peaks \noverlap with each ChromHMM state"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


p2 <- ggplot(final_table, aes(x = state_label, y = chromHMMstate_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(x = "ChromHMM state", y = "ChromHMM segments overlapped (%)",
       title = "Percentage of ChromHMM peaks \noverlap with OCR peaks") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p1)
print(p2)

```


## DNase peaks

```{r fig.height=4, fig.width=6}

peak_folder <- "/project/spott/kevinluo/Fiber_seq/data/DNaseseq/GM12878/"

peak_files <- list.files(
  path = peak_folder,
  pattern = "_peaks_.*\\.bed\\.gz$",
  full.names = TRUE
)
stopifnot(length(peak_files) >= 1)

other_list <- lapply(peak_files, fread, showProgress = FALSE)
peak_df <- rbindlist(other_list, use.names = FALSE, fill = TRUE)

gr_peaks <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3)
)

all_states <- sort(unique(ChromHMM_df$state))

min_bp <- 100L

## --- 1) One overlap for ALL states (fast) ---
hits <- findOverlaps(gr_peaks, gr_hmm, ignore.strand = TRUE)

# overlap width for each hit
ov_bp <- width(pintersect(gr_peaks[queryHits(hits)], gr_hmm[subjectHits(hits)]))

# keep only >= min_bp
keep <- ov_bp >= min_bp
qh <- queryHits(hits)[keep]      # peak indices
sh <- subjectHits(hits)[keep]    # chromhmm segment indices

# state label for each kept hit
st <- mcols(gr_hmm)$state[sh]

## --- 2) Summarize per state (unique peak hits, unique segment hits) ---
dt_hits <- data.table(
  state = st,
  peak_i = qh,
  seg_i  = sh
)

sum_by_state <- dt_hits[, .(
  OCRpeaks_overlap        = uniqueN(peak_i),
  chromHMMstate_overlap   = uniqueN(seg_i)
), by = state]

## --- 3) Add totals for each state and compute proportions ---
# totals
n_OCRpeaks <- length(gr_peaks)

dt_state_totals <- data.table(
  state = names(table(mcols(gr_hmm)$state)),
  n_chromHMMstate_peaks = as.integer(table(mcols(gr_hmm)$state))
)

summary_table <- merge(dt_state_totals, sum_by_state, by = "state", all.x = TRUE)

# states with no overlap -> NA to 0
for (col in c("OCRpeaks_overlap", "chromHMMstate_overlap")) {
  summary_table[is.na(get(col)), (col) := 0L]
}

summary_table[, `:=`(
  n_OCRpeaks = n_OCRpeaks,
  OCRpeaks_overlap_prop = OCRpeaks_overlap / n_OCRpeaks,
  chromHMMstate_overlap_prop = chromHMMstate_overlap / n_chromHMMstate_peaks
)]

## --- 4) Add annotation + select final columns ---
setDT(chromhmm_annot)

final_table <- merge(
  chromhmm_annot[, .(state, state_id, mnemonic, description)],
  summary_table[, .(
    state,
    n_OCRpeaks,
    n_chromHMMstate_peaks,
    OCRpeaks_overlap,
    OCRpeaks_overlap_prop,
    chromHMMstate_overlap,
    chromHMMstate_overlap_prop
  )],
  by = "state",
  all.x = TRUE
)

# if some states missing in summary_table (shouldn't), set to 0
for (col in c("n_OCRpeaks","n_chromHMMstate_peaks","OCRpeaks_overlap","chromHMMstate_overlap",
              "OCRpeaks_overlap_prop","chromHMMstate_overlap_prop")) {
  final_table[is.na(get(col)), (col) := if (grepl("prop$", col)) 0 else 0L]
}

final_table <- final_table[, .(
  state_id, mnemonic, description,
  n_OCRpeaks,
  n_chromHMMstate_peaks,
  OCRpeaks_overlap,
  OCRpeaks_overlap_prop,
  chromHMMstate_overlap,
  chromHMMstate_overlap_prop
)]

setorder(final_table, state_id)

DT::datatable(final_table,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Overlap summary'),options = list(pageLength = 10) )

final_table[, OCRpeaks_overlap_pct := OCRpeaks_overlap_prop * 100]
final_table[, chromHMMstate_overlap_pct := chromHMMstate_overlap_prop * 100]



final_table[, state_label := paste(state_id, mnemonic, sep = " ")]
final_table[, state_label := factor(state_label,
                                    levels = final_table[order(state_id), state_label])]

final_table <- merge(
  final_table,
  chromhmm_annot[, .(state_id, hex)],
  by = "state_id",
  all.x = TRUE
)

p1 <- ggplot(final_table,
       aes(x = state_label, y = OCRpeaks_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(
    x = "ChromHMM state",
    y = "OCR peaks overlapping (%)",
    title = "Percentage of OCR peaks \noverlap with each ChromHMM state"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


p2 <- ggplot(final_table, aes(x = state_label, y = chromHMMstate_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(x = "ChromHMM state", y = "ChromHMM segments overlapped (%)",
       title = "Percentage of ChromHMM peaks \noverlap with OCR peaks") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p1)
print(p2)

```

## ATAC-seq peaks -- IDR

```{r fig.height=4, fig.width=6}

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ATACseq/GM12878/ATACseq_GM12878_IDR_peaks_ENCFF429WTW.bed.gz",
  showProgress = FALSE
)

gr_peaks <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3)
)

all_states <- sort(unique(ChromHMM_df$state))

min_bp <- 100L

## --- 1) One overlap for ALL states (fast) ---
hits <- findOverlaps(gr_peaks, gr_hmm, ignore.strand = TRUE)

# overlap width for each hit
ov_bp <- width(pintersect(gr_peaks[queryHits(hits)], gr_hmm[subjectHits(hits)]))

# keep only >= min_bp
keep <- ov_bp >= min_bp
qh <- queryHits(hits)[keep]      # peak indices
sh <- subjectHits(hits)[keep]    # chromhmm segment indices

# state label for each kept hit
st <- mcols(gr_hmm)$state[sh]

## --- 2) Summarize per state (unique peak hits, unique segment hits) ---
dt_hits <- data.table(
  state = st,
  peak_i = qh,
  seg_i  = sh
)

sum_by_state <- dt_hits[, .(
  OCRpeaks_overlap        = uniqueN(peak_i),
  chromHMMstate_overlap   = uniqueN(seg_i)
), by = state]

## --- 3) Add totals for each state and compute proportions ---
# totals
n_OCRpeaks <- length(gr_peaks)

dt_state_totals <- data.table(
  state = names(table(mcols(gr_hmm)$state)),
  n_chromHMMstate_peaks = as.integer(table(mcols(gr_hmm)$state))
)

summary_table <- merge(dt_state_totals, sum_by_state, by = "state", all.x = TRUE)

# states with no overlap -> NA to 0
for (col in c("OCRpeaks_overlap", "chromHMMstate_overlap")) {
  summary_table[is.na(get(col)), (col) := 0L]
}

summary_table[, `:=`(
  n_OCRpeaks = n_OCRpeaks,
  OCRpeaks_overlap_prop = OCRpeaks_overlap / n_OCRpeaks,
  chromHMMstate_overlap_prop = chromHMMstate_overlap / n_chromHMMstate_peaks
)]

## --- 4) Add annotation + select final columns ---
setDT(chromhmm_annot)

final_table <- merge(
  chromhmm_annot[, .(state, state_id, mnemonic, description)],
  summary_table[, .(
    state,
    n_OCRpeaks,
    n_chromHMMstate_peaks,
    OCRpeaks_overlap,
    OCRpeaks_overlap_prop,
    chromHMMstate_overlap,
    chromHMMstate_overlap_prop
  )],
  by = "state",
  all.x = TRUE
)

# if some states missing in summary_table (shouldn't), set to 0
for (col in c("n_OCRpeaks","n_chromHMMstate_peaks","OCRpeaks_overlap","chromHMMstate_overlap",
              "OCRpeaks_overlap_prop","chromHMMstate_overlap_prop")) {
  final_table[is.na(get(col)), (col) := if (grepl("prop$", col)) 0 else 0L]
}

final_table <- final_table[, .(
  state_id, mnemonic, description,
  n_OCRpeaks,
  n_chromHMMstate_peaks,
  OCRpeaks_overlap,
  OCRpeaks_overlap_prop,
  chromHMMstate_overlap,
  chromHMMstate_overlap_prop
)]

setorder(final_table, state_id)

DT::datatable(final_table,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Overlap summary'),options = list(pageLength = 10) )

final_table[, OCRpeaks_overlap_pct := OCRpeaks_overlap_prop * 100]
final_table[, chromHMMstate_overlap_pct := chromHMMstate_overlap_prop * 100]



final_table[, state_label := paste(state_id, mnemonic, sep = " ")]
final_table[, state_label := factor(state_label,
                                    levels = final_table[order(state_id), state_label])]

final_table <- merge(
  final_table,
  chromhmm_annot[, .(state_id, hex)],
  by = "state_id",
  all.x = TRUE
)

p1 <- ggplot(final_table,
       aes(x = state_label, y = OCRpeaks_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(
    x = "ChromHMM state",
    y = "OCR peaks overlapping (%)",
    title = "Percentage of OCR peaks \noverlap with each ChromHMM state"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


p2 <- ggplot(final_table, aes(x = state_label, y = chromHMMstate_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(x = "ChromHMM state", y = "ChromHMM segments overlapped (%)",
       title = "Percentage of ChromHMM peaks \noverlap with OCR peaks") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p1)
print(p2)

```

## ATAC-seq peak -- pseudoreplicated

```{r fig.height=4, fig.width=6}

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ATACseq/GM12878/ATACseq_GM12878_pseudoreplicated_peaks_ENCFF748UZH.bed.gz",
  showProgress = FALSE
)

gr_peaks <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3)
)

all_states <- sort(unique(ChromHMM_df$state))

min_bp <- 100L

## --- 1) One overlap for ALL states (fast) ---
hits <- findOverlaps(gr_peaks, gr_hmm, ignore.strand = TRUE)

# overlap width for each hit
ov_bp <- width(pintersect(gr_peaks[queryHits(hits)], gr_hmm[subjectHits(hits)]))

# keep only >= min_bp
keep <- ov_bp >= min_bp
qh <- queryHits(hits)[keep]      # peak indices
sh <- subjectHits(hits)[keep]    # chromhmm segment indices

# state label for each kept hit
st <- mcols(gr_hmm)$state[sh]

## --- 2) Summarize per state (unique peak hits, unique segment hits) ---
dt_hits <- data.table(
  state = st,
  peak_i = qh,
  seg_i  = sh
)

sum_by_state <- dt_hits[, .(
  OCRpeaks_overlap        = uniqueN(peak_i),
  chromHMMstate_overlap   = uniqueN(seg_i)
), by = state]

## --- 3) Add totals for each state and compute proportions ---
# totals
n_OCRpeaks <- length(gr_peaks)

dt_state_totals <- data.table(
  state = names(table(mcols(gr_hmm)$state)),
  n_chromHMMstate_peaks = as.integer(table(mcols(gr_hmm)$state))
)

summary_table <- merge(dt_state_totals, sum_by_state, by = "state", all.x = TRUE)

# states with no overlap -> NA to 0
for (col in c("OCRpeaks_overlap", "chromHMMstate_overlap")) {
  summary_table[is.na(get(col)), (col) := 0L]
}

summary_table[, `:=`(
  n_OCRpeaks = n_OCRpeaks,
  OCRpeaks_overlap_prop = OCRpeaks_overlap / n_OCRpeaks,
  chromHMMstate_overlap_prop = chromHMMstate_overlap / n_chromHMMstate_peaks
)]

## --- 4) Add annotation + select final columns ---
setDT(chromhmm_annot)

final_table <- merge(
  chromhmm_annot[, .(state, state_id, mnemonic, description)],
  summary_table[, .(
    state,
    n_OCRpeaks,
    n_chromHMMstate_peaks,
    OCRpeaks_overlap,
    OCRpeaks_overlap_prop,
    chromHMMstate_overlap,
    chromHMMstate_overlap_prop
  )],
  by = "state",
  all.x = TRUE
)

# if some states missing in summary_table (shouldn't), set to 0
for (col in c("n_OCRpeaks","n_chromHMMstate_peaks","OCRpeaks_overlap","chromHMMstate_overlap",
              "OCRpeaks_overlap_prop","chromHMMstate_overlap_prop")) {
  final_table[is.na(get(col)), (col) := if (grepl("prop$", col)) 0 else 0L]
}

final_table <- final_table[, .(
  state_id, mnemonic, description,
  n_OCRpeaks,
  n_chromHMMstate_peaks,
  OCRpeaks_overlap,
  OCRpeaks_overlap_prop,
  chromHMMstate_overlap,
  chromHMMstate_overlap_prop
)]

setorder(final_table, state_id)

DT::datatable(final_table,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Overlap summary'),options = list(pageLength = 10) )

final_table[, OCRpeaks_overlap_pct := OCRpeaks_overlap_prop * 100]
final_table[, chromHMMstate_overlap_pct := chromHMMstate_overlap_prop * 100]



final_table[, state_label := paste(state_id, mnemonic, sep = " ")]
final_table[, state_label := factor(state_label,
                                    levels = final_table[order(state_id), state_label])]

final_table <- merge(
  final_table,
  chromhmm_annot[, .(state_id, hex)],
  by = "state_id",
  all.x = TRUE
)

p1 <- ggplot(final_table,
       aes(x = state_label, y = OCRpeaks_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(
    x = "ChromHMM state",
    y = "OCR peaks overlapping (%)",
    title = "Percentage of OCR peaks \noverlap with each ChromHMM state"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


p2 <- ggplot(final_table, aes(x = state_label, y = chromHMMstate_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(x = "ChromHMM state", y = "ChromHMM segments overlapped (%)",
       title = "Percentage of ChromHMM peaks \noverlap with OCR peaks") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p1)
print(p2)

```

## cCRE peak

```{r fig.height=4, fig.width=6}

peak_df <- fread(
  "/project/spott/kevinluo/Fiber_seq/data/ENCODE_CREs/cCREs_v4/LCLs/GM12878_cCREs_ENCFF428XFI_ENCFF280PUF_ENCFF469WVA_ENCFF644EEX.bed",
  showProgress = FALSE
)
peak_df <- peak_df[peak_df$V10 !="Low-DNase",]

gr_peaks <- GRanges(
  seqnames = peak_df$V1,
  ranges   = IRanges(start = peak_df$V2 + 1L, end = peak_df$V3)
)

all_states <- sort(unique(ChromHMM_df$state))

min_bp <- 100L

## --- 1) One overlap for ALL states (fast) ---
hits <- findOverlaps(gr_peaks, gr_hmm, ignore.strand = TRUE)

# overlap width for each hit
ov_bp <- width(pintersect(gr_peaks[queryHits(hits)], gr_hmm[subjectHits(hits)]))

# keep only >= min_bp
keep <- ov_bp >= min_bp
qh <- queryHits(hits)[keep]      # peak indices
sh <- subjectHits(hits)[keep]    # chromhmm segment indices

# state label for each kept hit
st <- mcols(gr_hmm)$state[sh]

## --- 2) Summarize per state (unique peak hits, unique segment hits) ---
dt_hits <- data.table(
  state = st,
  peak_i = qh,
  seg_i  = sh
)

sum_by_state <- dt_hits[, .(
  OCRpeaks_overlap        = uniqueN(peak_i),
  chromHMMstate_overlap   = uniqueN(seg_i)
), by = state]

## --- 3) Add totals for each state and compute proportions ---
# totals
n_OCRpeaks <- length(gr_peaks)

dt_state_totals <- data.table(
  state = names(table(mcols(gr_hmm)$state)),
  n_chromHMMstate_peaks = as.integer(table(mcols(gr_hmm)$state))
)

summary_table <- merge(dt_state_totals, sum_by_state, by = "state", all.x = TRUE)

# states with no overlap -> NA to 0
for (col in c("OCRpeaks_overlap", "chromHMMstate_overlap")) {
  summary_table[is.na(get(col)), (col) := 0L]
}

summary_table[, `:=`(
  n_OCRpeaks = n_OCRpeaks,
  OCRpeaks_overlap_prop = OCRpeaks_overlap / n_OCRpeaks,
  chromHMMstate_overlap_prop = chromHMMstate_overlap / n_chromHMMstate_peaks
)]

## --- 4) Add annotation + select final columns ---
setDT(chromhmm_annot)

final_table <- merge(
  chromhmm_annot[, .(state, state_id, mnemonic, description)],
  summary_table[, .(
    state,
    n_OCRpeaks,
    n_chromHMMstate_peaks,
    OCRpeaks_overlap,
    OCRpeaks_overlap_prop,
    chromHMMstate_overlap,
    chromHMMstate_overlap_prop
  )],
  by = "state",
  all.x = TRUE
)

# if some states missing in summary_table (shouldn't), set to 0
for (col in c("n_OCRpeaks","n_chromHMMstate_peaks","OCRpeaks_overlap","chromHMMstate_overlap",
              "OCRpeaks_overlap_prop","chromHMMstate_overlap_prop")) {
  final_table[is.na(get(col)), (col) := if (grepl("prop$", col)) 0 else 0L]
}

final_table <- final_table[, .(
  state_id, mnemonic, description,
  n_OCRpeaks,
  n_chromHMMstate_peaks,
  OCRpeaks_overlap,
  OCRpeaks_overlap_prop,
  chromHMMstate_overlap,
  chromHMMstate_overlap_prop
)]

setorder(final_table, state_id)

DT::datatable(final_table,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Overlap summary'),options = list(pageLength = 10) )

final_table[, OCRpeaks_overlap_pct := OCRpeaks_overlap_prop * 100]
final_table[, chromHMMstate_overlap_pct := chromHMMstate_overlap_prop * 100]



final_table[, state_label := paste(state_id, mnemonic, sep = " ")]
final_table[, state_label := factor(state_label,
                                    levels = final_table[order(state_id), state_label])]

final_table <- merge(
  final_table,
  chromhmm_annot[, .(state_id, hex)],
  by = "state_id",
  all.x = TRUE
)

p1 <- ggplot(final_table,
       aes(x = state_label, y = OCRpeaks_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(
    x = "ChromHMM state",
    y = "OCR peaks overlapping (%)",
    title = "Percentage of OCR peaks \noverlap with each ChromHMM state"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


p2 <- ggplot(final_table, aes(x = state_label, y = chromHMMstate_overlap_pct)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  theme_classic(base_size = 14) +
  labs(x = "ChromHMM state", y = "ChromHMM segments overlapped (%)",
       title = "Percentage of ChromHMM peaks \noverlap with OCR peaks") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p1)
print(p2)

```