---
title: "TF Co-accessibility for NRF1 & SP1, motif sites are 400 - 800 bp apart"
author: "XSun"
date: "2025-08-31"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---


```{r warning=F, message=F}
library(data.table)
library(reshape2)
library(GenomicRanges)
library(ggplot2)
library(tidyverse)

source("/project/spott/kevinluo/Fiber_seq/fiberhub/code/process_fiberseq_data.R")
source("/project/spott/kevinluo/Fiber_seq/fiberhub/code/plots.R")

TF1 <- "NRF1"
TF2 <- "SP1"
```

# Extract the regions containing both TF motif sites

Regions containing both TFs were extracted, and only those in which the TFs were separated by more than `400 bp` but less than `800 bp` were retained. Each motif was then extended by 500 bp on both sides to define the final regions.


```{r warning=F, message=F}

motif_sites_motif1 <- readRDS(paste0('/project/spott/kevinluo/Fiber_seq/processed_data/hg38/TFs/',TF1,'.GM12878.sites.chip.labels.rds'))
motif_sites_motif1_chip1 <- motif_sites_motif1[motif_sites_motif1$chip_label == 1,]

motif_sites_motif2 <- readRDS(paste0('/project/spott/kevinluo/Fiber_seq/processed_data/hg38/TFs/',TF2,'.GM12878.sites.chip.labels.rds'))
motif_sites_motif2_chip1 <- motif_sites_motif1[motif_sites_motif2$chip_label == 1,]

motif1 <- motif_sites_motif1_chip1
motif1 <- motif1[!is.na(motif1$start),]
motif2 <- motif_sites_motif2_chip1
motif2 <- motif2[!is.na(motif2$start),]

## Convert to GRanges
gr_motif1 <- GRanges(seqnames = motif1$chr,
                 ranges   = IRanges(start = motif1$start+100, end = motif1$end-100),
                 site     = motif1$name)

gr_motif2 <- GRanges(seqnames = motif2$chr,
                 ranges   = IRanges(start = motif2$start+100, end = motif2$end-100),
                 site     = motif2$name)

## Define distance cutoff
dist_cutoff_large <- 800
dist_cutoff_small <- 400

## Find nearest motif2 site for each motif1 site
nearest_hits <- distanceToNearest(gr_motif1, gr_motif2)

## Filter by distance threshold
close_pairs <- nearest_hits[mcols(nearest_hits)$distance <= dist_cutoff_large & mcols(nearest_hits)$distance >= dist_cutoff_small]

## Extract regions
motif1_close <- gr_motif1[queryHits(close_pairs)]
motif2_close <- gr_motif2[subjectHits(close_pairs)]

## Make a data.frame
df_close <- data.frame(
  chr = as.character(seqnames(motif1_close)),
  motif1_start = start(motif1_close),
  motif1_end   = end(motif1_close),
  motif1_site  = motif1_close$site,
  motif2_start  = start(motif2_close),
  motif2_end    = end(motif2_close),
  motif2_site   = motif2_close$site,
  distance   = mcols(close_pairs)$distance
)

DT::datatable(df_close,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Motif sites selected'),options = list(pageLength = 5) )

df_summary <- df_close %>%
  group_by(chr, motif2_site, motif2_start, motif2_end) %>%
  summarise(
    min_motif1_start = min(motif1_start),
    max_motif1_end   = max(motif1_end),
    .groups = "drop"
  )

df_final <- df_summary %>%
  transmute(
    chr,
    start = pmin(motif2_start, min_motif1_start) - 500,
    end   = pmax(motif2_end,   max_motif1_end)   + 500
  )

df_final <- df_final %>%
  mutate(chr_num = as.numeric(gsub("chr", "", chr))) %>%
  arrange(chr_num, start) %>%
  select(-chr_num)

DT::datatable(df_final,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Regions selected for plotting '),options = list(pageLength = 5) )

```

# Visualization and co-accessibility test 

For visualization, highlighted windows are the ones `±7 bp` around each motif. 

To determine whether a TF has a footprint, footprint data were extracted from `res$fiberHMM_fps_infire`. 

Footprints were retained if they (1) had a FiberHMM footprint size <40 bp and (2) overlapped with the highlighted motif windows. 

Next, reads covering both motif regions were extracted, and the presence or absence of footprints for the two TFs was counted.

Last, we did FET


```{r message=FALSE,warning=FALSE, fig.height=10, fig.width=10}

res <- readRDS("/project/xinhe/xsun/website/fiberseq/data/example_res_from_plot_region_reads.RDS")
print(res$fiberHMM_fps_infire)
head(res$fiberHMM_fps_infire)

```




```{r message=FALSE,warning=FALSE, fig.height=10, fig.width=10}

out_dir="/project/xinhe/xsun/fiberseq/1.TFs/plot_region/results"
sample_name="LCL_18508"

df_close_nodup <- df_close[!duplicated(df_close$motif2_site),]

#for (i in 1:3){
for (i in 1:nrow(df_final)){
  
  test_range_str <- paste0(df_final$chr[i],":",df_final$start[i],"-",df_final$end[i])
  plot_name=paste0(TF1,"_",TF2,"_",sample_name, "_", test_range_str)
  result_dir = paste0(out_dir, "/", plot_name, "/parsed")
  
  region <- get_region(result_dir)
  subtitle <- sprintf("%s:%s-%s", region$chr, region$start, region$end)
  
  print(subtitle)
  
  df_close_region <- df_close[df_close$motif2_site == df_close_nodup$motif2_site[i],]
  min_start_motif1 <- min(df_close_region$motif1_start)
  max_end_motif1   <- max(df_close_region$motif1_end)
  
  highlight_windows.df <- data.frame(start = c(min_start_motif1 - 7,df_close_nodup$motif2_start[i] -7 ), 
                              end = c(max_end_motif1 + 7 ,df_close_nodup$motif2_end[i] +7))
  
  plot_name=paste0(TF1,"_",TF2,"_",sample_name, "_", test_range_str)
  result_dir = paste0(out_dir, "/", plot_name, "/parsed")

  
  res <- plot_region_reads(result_dir, 
                         filter_fire_reads = FALSE,
                         cover_region_only = FALSE,
                         cluster_reads = TRUE,
                         select_A_only = TRUE,
                         split_haplotype = FALSE,
                         highlight_window = highlight_windows.df,
                         color_highlight_window = "#ade6b3",
                         plot_name = plot_name,
                         subtitle = subtitle)

  region <- res$region
  window_size <- res$window_size
  rids_df <- res$rids_df
  pileup <- res$pileup
  reads <- res$reads
  fire <- res$fire
  fdr_fire <- res$fdr_fire
  fiberHMM_fps <- res$fiberHMM_fps
  gg <- res$gg
  
  
  print(cowplot::plot_grid(
    gg$pileup,
    gg$dimelo_reads,
    gg$fire_fdr,
    gg$fire_reads,
    gg$fiberHMM,
    ncol = 1,
    rel_heights = c(4,8,2,8,8),
    align = "v",
    axis = "b"
  ))

  
  fps <- res$fiberHMM_fps_infire
  fps_rid <- fps$RID
  
  #fps_highlight_motif1 <- fps[fps$start > highlight_windows.df$start[1] & fps$end < highlight_windows.df$end[1], ]
  fps_highlight_motif1 <- fps[fps$start <= highlight_windows.df$end[1] & fps$end >= highlight_windows.df$start[1], ]
  fps_tf_motif1 <- fps_highlight_motif1[fps_highlight_motif1$class == "< 40bp",]
  rid_motif1 <-   as.character(res$rids_df$RID[res$rids_df$start <= max_end_motif1 & res$rids_df$end >= min_start_motif1])
  
  
  #fps_highlight_motif2 <- fps[fps$start > highlight_windows.df$start[2] & fps$end < highlight_windows.df$end[2], ]
  fps_highlight_motif2 <- fps[fps$start <= highlight_windows.df$end[2] & fps$end >= highlight_windows.df$start[2], ]
  fps_tf_motif2 <- fps_highlight_motif2[fps_highlight_motif2$class == "< 40bp",]
  rid_motif2 <-   as.character(res$rids_df$RID[res$rids_df$start <= df_close_nodup$motif2_end[i] & res$rids_df$end >= df_close_nodup$motif2_start[i]])
  
  RIDs_overlap_both_regions <- intersect(rid_motif1, rid_motif2)
  
  RIDs_access.df <- data.frame(RID = RIDs_overlap_both_regions, 
                               TF1 = FALSE,
                               TF2 = FALSE)
  
  RIDs_access.df$TF1[RIDs_access.df$RID %in% unique(fps_tf_motif1$RID)] <- TRUE
  RIDs_access.df$TF2[RIDs_access.df$RID %in% unique(fps_tf_motif2$RID)] <- TRUE
  
  
  # print(RIDs_access.df)
  
  if(!(sum(RIDs_access.df$TF1) > 0 & sum(RIDs_access.df$TF2) > 0)) next 
  
  cat("Create a 2x2 table of coaccessibility and run fisher's exact test...\n")
  print(contingency_table <- table(RIDs_access.df[,c("TF1", "TF2")]))
  
  print(fisher_res <- fisher.test(contingency_table))
  
}




```







