---
title: "TF Co-accessibility for NRF1 pairs"
author: "XSun"
date: "2025-09-03"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---


```{r warning=F, message=F}
library(data.table)
library(reshape2)
library(GenomicRanges)
library(ggplot2)
library(tidyverse)
library(gridExtra)

source("/project/spott/kevinluo/Fiber_seq/fiberhub/code/process_fiberseq_data.R")
source("/project/spott/kevinluo/Fiber_seq/fiberhub/code/plots.R")

TF1 <- "NRF1"
```


# Extract the regions containing TF pairs

TF pairs for one TF is two separate recognition motifs of one TF within the same CRE (<140 bp). Ref: https://www.cell.com/molecular-cell/fulltext/S1097-2765(20)30793-0?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1097276520307930%3Fshowall%3Dtrue

We extracted regions containing both motifs of a TF pair, retaining only those in which the motifs were separated by more than 14 bp but less than 140 bp.


The detailed procedure is: 

1. Load the motif sites and filter for `chip_label == 1`.

```{r warning=F, message=F}


# Read motif sites and filter
motif_sites <- readRDS(
  file.path("/project/spott/kevinluo/Fiber_seq/processed_data/hg38/TFs",
            paste0(TF1, ".GM12878.sites.chip.labels.rds"))
) %>%
  filter(chip_label == 1, !is.na(start))


DT::datatable(motif_sites,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Motif sites'),options = list(pageLength = 5) )
```


2. Identify all possible pairs of motifs for the TF on the same chromosome, excluding self-pairs.

3. Calculate the genomic distance between motif pairs and retain only those separated by `14 - 140bp`.


```{r warning=F, message=F}
# Convert to GRanges
gr_motif <- GRanges(
  seqnames = motif_sites$chr,
  ranges   = IRanges(start = motif_sites$start + 100,
                     end   = motif_sites$end   - 100),
  site     = motif_sites$name
)

# Distance cutoffs
dist_cutoff_large <- 140L
dist_cutoff_small <- 15L

# Find all motif pairs within large distance
hits <- findOverlaps(gr_motif, gr_motif,
                     ignore.strand = TRUE,
                     maxgap = dist_cutoff_large)

q <- queryHits(hits)
s <- subjectHits(hits)

# Keep unique orientation, no self-pairs
keep <- q < s
q <- q[keep]; s <- s[keep]

# Distances
d <- distance(gr_motif[q], gr_motif[s])

# Filter by [small, large]
sel <- d >= dist_cutoff_small & d <= dist_cutoff_large
q <- q[sel]; s <- s[sel]; d <- d[sel]

# Build pair dataframe
df_close <- data.frame(
  chr          = as.character(seqnames(gr_motif[q])),
  motif1_start = start(gr_motif[q]),
  motif1_end   = end(gr_motif[q]),
  motif1_site  = gr_motif[q]$site,
  motif2_start = start(gr_motif[s]),
  motif2_end   = end(gr_motif[s]),
  motif2_site  = gr_motif[s]$site,
  distance     = as.integer(d)
)

df_close$id <- paste0(df_close$motif1_site,"-", df_close$motif2_site)

DT::datatable(df_close,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Motif pairs kept'),options = list(pageLength = 5) )




```

4. For regions with multiple pairs, define broader genomic intervals covering both motifs in each pair, adding a 500 bp buffer on each side.

5. Round these intervals to the nearest hundred bases to obtain the final regions.


```{r warning=F, message=F}
# Summarize per motif2 and build final regions
df_final <- df_close %>%
  group_by(chr, motif2_site, motif2_start, motif2_end) %>%
  mutate(
    min_motif1_start = min(motif1_start),
    max_motif1_end   = max(motif1_end)
  ) %>%
  ungroup() %>%
  transmute(
    chr,
    start = pmin(motif2_start, min_motif1_start) - 500,
    end   = pmax(motif2_end,   max_motif1_end)   + 500
  ) %>%
  mutate(
    chr_num = as.numeric(gsub("chr", "", chr)),
    start   = floor(start / 100) * 100,
    end     = ceiling(end / 100) * 100
  ) %>%
  arrange(chr_num, start) %>%
  select(-chr_num) 

df_final$id <- paste0(df_final$chr,"-", df_final$start, "-",df_final$end)

df_final_region <- df_final[!duplicated(df_final$id),]

DT::datatable(df_final_region,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Regions covering the motif pairs'),options = list(pageLength = 5) )

```


# Visualization and co-accessibility test 

For visualization, highlighted windows are the ones `±7 bp` around each motif. 

To determine whether a TF has a footprint, footprint data were extracted from `res$fiberHMM_fps_infire`. 

Footprints were retained if they (1) had a FiberHMM footprint size <40 bp and (2) overlapped with the highlighted motif windows. 

Next, reads covering both motif regions were extracted, and the presence or absence of footprints for the two TFs was counted.

Last, we did FET 

## Co-occupied rate

```{r message=FALSE,warning=FALSE, fig.height=10, fig.width=10}


FET_or <- readRDS("/project/xinhe/xsun/fiberseq/1.TFs/results/NRF1_pairs/tmp/FET_or.RDS")
FETp <- readRDS("/project/xinhe/xsun/fiberseq/1.TFs/results/NRF1_pairs/tmp/FETp.RDS")
#plots <- readRDS("/project/xinhe/xsun/fiberseq/1.TFs/results/NRF1_pairs/plots_naremoved2.RDS")
FET_matrix <- readRDS("/project/xinhe/xsun/fiberseq/1.TFs/results/NRF1_pairs/tmp/FET_matrix.RDS")
FET_res <- readRDS("/project/xinhe/xsun/fiberseq/1.TFs/results/NRF1_pairs/tmp/FET_res.RDS")

FET_res_all <- data.frame(pair_region = names(FETp), FET_pvalues = FETp, FET_oddratio = FET_or)
FET_res_all$fdr <- p.adjust(FET_res_all$FET_pvalues, method = "fdr")

df_close$order <- seq_len(nrow(df_close))  # add index column
df_close <- merge(
  df_close, FET_res_all,
  by.x = "id", by.y = "pair_region",
  all.x = TRUE, sort = FALSE
)
df_close <- df_close[order(df_close$order), ]  # restore original order
df_close$order <- NULL  # drop helper column

df_close$FET_pvalues <- round(df_close$FET_pvalues, digits = 4)
df_close$FET_oddratio <- round(df_close$FET_oddratio, digits = 4)
df_close$fdr <- round(df_close$fdr, digits = 4)

print(paste0("total pairs = ", nrow(FET_res_all)))


DT::datatable(df_close,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','FET results for the TF pairs'),options = list(pageLength = 5) )


```


Using `FDR < 0.1` as cutoff, 

The co-occupancy rate for the pairs: 

```{r message=FALSE,warning=FALSE, fig.height=4, fig.width=10}

# keep only rows with distance and fdr
df_filtered <- df_close %>%
  filter(!is.na(fdr), !is.na(distance)) %>%
  # restrict to the 14-140 bp range so the overall row is the sum of the three sub-bins
  filter(distance >= 14, distance <= 140)

# create non-overlapping bins
df_binned <- df_filtered %>%
  mutate(bin = case_when(
    distance >= 14  & distance <= 50  ~ "14-50bp",
    distance >  50  & distance <= 70  ~ "50-70bp",
    distance >  70  & distance <= 140 ~ "70-140bp",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(bin))

# summary for the three sub-bins
rates_df <- df_binned %>%
  group_by(bin) %>%
  summarise(
    total_pairs = n(),
    sig_pairs   = sum(fdr < 0.1),
    cooccupancy_rate = ifelse(total_pairs > 0, sig_pairs / total_pairs, NA_real_)
  ) %>%
  ungroup()

# add overall (14-140bp) as the sum of the three sub-bins
overall <- rates_df %>%
  summarise(
    total_pairs = sum(total_pairs),
    sig_pairs   = sum(sig_pairs)
  ) %>%
  mutate(
    bin = "14-140bp",
    cooccupancy_rate = ifelse(total_pairs > 0, sig_pairs / total_pairs, NA_real_)
  ) %>%
  select(bin, total_pairs, sig_pairs, cooccupancy_rate)

rates_df <- bind_rows(overall, rates_df) %>%
  # set plotting order: overall first, then sub-bins
  mutate(bin = factor(bin, levels = c("14-140bp", "14-50bp", "50-70bp", "70-140bp")))

print(rates_df)

# prepare stacked counts: significant vs non-significant (so bars sum to total_pairs)
counts_df <- rates_df %>%
  mutate(nonsig_pairs = total_pairs - sig_pairs) %>%
  select(bin, sig_pairs, nonsig_pairs) %>%
  pivot_longer(cols = c("nonsig_pairs", "sig_pairs"),
               names_to = "pair_type",
               values_to = "count") %>%
  mutate(
    pair_type = factor(pair_type, levels = c("nonsig_pairs", "sig_pairs"),
                       labels = c("Other pairs", "Co-occupied pairs \n(FDR < 0.1)"))
  )

# stacked barplot (total bar = total_pairs; blue segment = significant pairs)
p1 <- ggplot(counts_df, aes(x = bin, y = count, fill = pair_type)) +
  geom_col() +
  # counts inside each segment
  geom_text(aes(label = count), position = position_stack(vjust = 0.5), size = 3.5, color = "white") +
  # total count on top of each bar
  geom_text(
    data = rates_df,
    aes(x = bin, y = total_pairs, label = total_pairs),
    vjust = -0.5, inherit.aes = FALSE, size = 3.8
  ) +
  scale_fill_manual(values = c("steelblue", "orange")) +
  theme_minimal(base_size = 12) +
  labs(
    x = "Distance bin",
    y = "Number of pairs",
    fill = "",
    title = "Co-occupied pairs\n(FDR < 0.1)"
  ) +
  coord_cartesian(ylim = c(0, max(rates_df$total_pairs, na.rm = TRUE) * 1.08))


p2 <- ggplot(rates_df, aes(x = bin, y = cooccupancy_rate)) +
  geom_col(fill = "orange") +
  geom_text(aes(label = round(cooccupancy_rate, 3)), 
            vjust = -0.5, size = 5) +
  ylim(0, 0.5) +
  theme_minimal(base_size = 12) +
  labs(
    x = "Distance bin",
    y = "Co-occupancy rate",
    title = "Co-occupancy rates \n(FDR < 0.1)"
  )

grid.arrange(p1, p2, ncol = 2)
```


Using `FET p-value < 0.05` as cutoff, 

The co-occupancy rate for the pairs: 


```{r message=FALSE,warning=FALSE, fig.height=4, fig.width=10}

# summary for the three sub-bins
rates_df <- df_binned %>%
  group_by(bin) %>%
  summarise(
    total_pairs = n(),
    sig_pairs   = sum(FET_pvalues < 0.05),
    cooccupancy_rate = ifelse(total_pairs > 0, sig_pairs / total_pairs, NA_real_)
  ) %>%
  ungroup()

# add overall (14-140bp) as the sum of the three sub-bins
overall <- rates_df %>%
  summarise(
    total_pairs = sum(total_pairs),
    sig_pairs   = sum(sig_pairs)
  ) %>%
  mutate(
    bin = "14-140bp",
    cooccupancy_rate = ifelse(total_pairs > 0, sig_pairs / total_pairs, NA_real_)
  ) %>%
  select(bin, total_pairs, sig_pairs, cooccupancy_rate)

rates_df <- bind_rows(overall, rates_df) %>%
  # set plotting order: overall first, then sub-bins
  mutate(bin = factor(bin, levels = c("14-140bp", "14-50bp", "50-70bp", "70-140bp")))

print(rates_df)

# prepare stacked counts: significant vs non-significant (so bars sum to total_pairs)
counts_df <- rates_df %>%
  mutate(nonsig_pairs = total_pairs - sig_pairs) %>%
  select(bin, sig_pairs, nonsig_pairs) %>%
  pivot_longer(cols = c("nonsig_pairs", "sig_pairs"),
               names_to = "pair_type",
               values_to = "count") %>%
  mutate(
    pair_type = factor(pair_type, levels = c("nonsig_pairs", "sig_pairs"),
                       labels = c("Other pairs", "Co-occupied pairs \n(FET_pvalues < 0.05)"))
  )

# stacked barplot (total bar = total_pairs; blue segment = significant pairs)
p1 <- ggplot(counts_df, aes(x = bin, y = count, fill = pair_type)) +
  geom_col() +
  # counts inside each segment
  geom_text(aes(label = count), position = position_stack(vjust = 0.5), size = 3.5, color = "white") +
  # total count on top of each bar
  geom_text(
    data = rates_df,
    aes(x = bin, y = total_pairs, label = total_pairs),
    vjust = -0.5, inherit.aes = FALSE, size = 3.8
  ) +
  scale_fill_manual(values = c("steelblue", "orange")) +
  theme_minimal(base_size = 12) +
  labs(
    x = "Distance bin",
    y = "Number of pairs",
    fill = "",
    title = "Co-occupied pairs\n (FET_pvalues < 0.05)"
  ) +
  coord_cartesian(ylim = c(0, max(rates_df$total_pairs, na.rm = TRUE) * 1.08))


p2 <- ggplot(rates_df, aes(x = bin, y = cooccupancy_rate)) +
  geom_col(fill = "orange") +
  geom_text(aes(label = round(cooccupancy_rate, 3)), 
            vjust = -0.5, size = 5) +
  ylim(0, 0.5) +
  theme_minimal(base_size = 12) +
  labs(
    x = "Distance bin",
    y = "Co-occupancy rate",
    title = "Co-occupancy rates \n(FET_pvalues < 0.05)"
  )

grid.arrange(p1, p2, ncol = 2)
```

## Visualization

We randomly visualize some regions with co-occupied motifs (FDR < 0.1).


```{r message=FALSE,warning=FALSE, fig.height=10, fig.width=10}

index_sig <- which(df_close$fdr < 0.05)

set.seed(1)
index_sig <- index_sig[sample(x = 1:length(index_sig),size = 15,replace = F)]

out_dir="/project/xinhe/xsun/fiberseq/1.TFs/plot_region/results/NRF1/"
sample_name="LCL_18508"

for (i in 1:length(index_sig)){

  sig_pair <- df_close$id[index_sig[i]]

  index <- which(df_close$id == sig_pair)


  test_range_str <- paste0(df_final$chr[index],":",df_final$start[index],"-",df_final$end[index])
  plot_name=paste0(TF1,"_pairs_",sample_name, "_", test_range_str)
  result_dir = paste0(out_dir, "/", plot_name, "/parsed")

  region <- get_region(result_dir)
  subtitle <- sprintf("%s:%s-%s", region$chr, region$start, region$end)

  print(subtitle)
  print(sig_pair)

  df_close_region <- df_close[index,]
  start_motif1 <- df_close_region$motif1_start
  end_motif1   <- df_close_region$motif1_end

  start_motif2 <- df_close_region$motif2_start
  end_motif2   <- df_close_region$motif2_end

  highlight_windows.df <- data.frame(start = c(start_motif1 - 7, start_motif2 -7 ),
                                     end = c(end_motif1 + 7, end_motif2 +7))


  res <- plot_region_reads(result_dir,
                           filter_fire_reads = FALSE,
                           cover_region_only = FALSE,
                           cluster_reads = TRUE,
                           select_A_only = TRUE,
                           split_haplotype = FALSE,
                           highlight_window = highlight_windows.df,
                           color_highlight_window = "#ade6b3",
                           plot_name = plot_name,
                           subtitle = subtitle)

  gg <- res$gg

  print(cowplot::plot_grid(
    gg$pileup,
    gg$dimelo_reads,
    gg$fire_fdr,
    gg$fire_reads,
    gg$fiberHMM,
    ncol = 1,
    rel_heights = c(4,8,2,8,8),
    align = "v",
    axis = "b"
  ))

  print(FET_matrix[[index]])
  print(FET_res[[index]])

}



# plots_select <- plots[index_sig]
# FET_matrix_select <- FET_matrix[index_sig]
# FET_res_select <- FET_res[index_sig]
# for (i in 1:length(plots_select)){
#   
#   gg <- plots_select[[i]]
#   
#   print(cowplot::plot_grid(
#     gg$pileup,
#     gg$dimelo_reads,
#     gg$fire_fdr,
#     gg$fire_reads,
#     gg$fiberHMM,
#     ncol = 1,
#     rel_heights = c(4,8,2,8,8),
#     align = "v",
#     axis = "b"
#   ))
#   
#   print(FET_matrix_select[[i]])
#   print(FET_res_select[[i]])
#   
# 
# }

```






