---
title: "Validating the footprints using DNase footprint -- refined peaks"
author: "XSun"
date: "2026-03-16"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r warning=F, message=F}
library(data.table)
library(GenomicRanges)
library(ggplot2)
library(patchwork)

```


# Introduction

To validate footprints, we used DNase footprints from: https://www.vierstra.org/resources/dgf

They have two LCL lines: https://resources.altius.org/~jvierstra/projects/footprinting.2020/Supplementary_Table_1.xlsx

We used the GM12865: https://resources.altius.org/~jvierstra/projects/footprinting.2020/per.dataset/GM12865-DS12436/

DNase footprint file: `/project/spott/kevinluo/Fiber_seq/data/Vierstra_DGF/interval.all.fps.0.05.bed.gz`

Our footprint files are `/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_peaks/refined_peaks/`

- 10–30 bp
- 30–60 bp
- 60–80 bp

For each footprint set, we varied the footprint significance threshold q value (0.01, 0.1) and measured how much footprint sequence overlaps DNase peaks on **base-pair level**.



# Results

```{r}
fp_lengths <- c("10-30bp", "30-60bp", "60-80bp")

folder_fp <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_peaks/refined_peaks/"

## -----------------------------
## DNase peaks: read once
## -----------------------------
file_DNase_peak <- "/project/spott/kevinluo/Fiber_seq/data/Vierstra_DGF/interval.all.fps.0.05.bed.gz"
DNase_peak <- fread(file_DNase_peak, showProgress = FALSE)
setnames(DNase_peak, c(
  "dnase_chr","dnase_start","dnase_end","dnase_strand","dnase_pval"
))

setDT(DNase_peak)

gr_dnase <- GRanges(
  seqnames = DNase_peak$dnase_chr,
  ranges = IRanges(
    start = DNase_peak$dnase_start + 1,   ## 0-based -> 1-based
    end   = DNase_peak$dnase_end
  ),
  #strand = DNase_peak$dnase_strand
)

dnase_total_bp <- sum(width(reduce(gr_dnase)))
```


## q-values

```{r fig.width=13, fig.height=4}
q_cutoffs <- c(0.1, 0.01)

all_res <- list()
k <- 1

for (fp_length in fp_lengths) {
  
  res_list <- vector("list", length(q_cutoffs))
  for (i in seq(q_cutoffs)){
    
    
    ## -----------------------------
    ## read footprint file
    ## -----------------------------
    fp_file <- file.path(
      folder_fp,
      paste0("combined_all_chrs_", fp_length, "_macs3_q", q_cutoffs[i], "_minratio0.5_refined_peaks.bed.gz")
    )
    
    df_fp <- fread(fp_file, header = FALSE, showProgress = FALSE)
    
    setnames(df_fp, c(
      "fp_chr","fp_start","fp_end"
    ))
    
    setDT(df_fp)
    
    gr_fp <- GRanges(
      seqnames = df_fp$fp_chr,
      ranges = IRanges(
        start = df_fp$fp_start + 1,   ## 0-based -> 1-based
        end   = df_fp$fp_end
      )
    )
    
    fp_total_bp <- sum(width(reduce(gr_fp)))
    
    ## overlap
    hits <- findOverlaps(gr_fp, gr_dnase, ignore.strand = TRUE)
    
    if (length(hits) == 0) {
      overlap_bp <- 0
    } else {
      overlap_ranges <- pintersect(
        gr_fp[queryHits(hits)],
        gr_dnase[subjectHits(hits)]
      )
      overlap_bp <- sum(width(reduce(overlap_ranges)))
    }
    
    res_list[[i]] <- data.table(
      fp_length = fp_length,
      fp_qValue_cutoff = q_cutoffs[i],
      n_fp = length(gr_fp),
      fp_total_bp = fp_total_bp,
      overlap_bp = overlap_bp,
      prop_fp_bp_overlap_DNase = overlap_bp / fp_total_bp,
      prop_DNase_bp_overlap_FP = overlap_bp / dnase_total_bp
    )
    
    
    
    
  }
  
  
  all_res[[k]] <- rbindlist(res_list)
  k <- k + 1
    
}
  
overlap_res_all <- rbindlist(all_res)


library(ggplot2)
library(patchwork)

## ------------------------------
## Plot 1: number of footprints
## ------------------------------
p_nfp <- ggplot(
  overlap_res_all,
  aes(
    x = factor(fp_qValue_cutoff),
    y = n_fp,
    color = fp_length,
    group = fp_length
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  theme_classic(base_size = 14) +
  labs(
    x = "qValue cutoff",
    y = "Number of footprints(log 10 scale)",
    color = "Footprint length"
  )

## ------------------------------
## Plot 2: FP bp overlapping DNase
## ------------------------------
p1 <- ggplot(
  overlap_res_all,
  aes(
    x = factor(fp_qValue_cutoff),
    y = prop_fp_bp_overlap_DNase,
    color = fp_length,
    group = fp_length
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_classic(base_size = 14) +
  labs(
    x = "qValue cutoff",
    y = "Proportion of FP bp \noverlapping DNase",
    color = "Footprint length"
  )

## ------------------------------
## Plot 3: DNase bp overlapped by FP
## ------------------------------
p2 <- ggplot(
  overlap_res_all,
  aes(
    x = factor(fp_qValue_cutoff),
    y = prop_DNase_bp_overlap_FP,
    color = fp_length,
    group = fp_length
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_classic(base_size = 14) +
  labs(
    x = "qValue cutoff",
    y = "Proportion of DNase bp \noverlapped by FP",
    color = "Footprint length"
  )

## ------------------------------
## Combine plots
## ------------------------------
p_nfp + p1 + p2



```
